---
output: html_document
editor_options: 
  chunk_output_type: console
---

{{< include _setup.qmd >}}

# DOB PMT revenue option 2025-04-28

-   July 1, 2025
-   SLG employers in suburbs eliminate PMT
-   private employers \> \$10m +29.5bp
-   1.25-1.5 million cut rate from 11bp to 5.5bp
-   1.5m to 1.75m from 23bp to 11.5bp
-   raise self-employment threshold \$50k to \$150k

points of reference:

-   DOB CY 2025 FULL YEAR \$1.37 billion
-   DOB CY 2026 \$1.443 billion
-   KW: DOB about 10% higher per basis point
-   KW: DOB private 1 bp \> 10m private \$40.75 bp in 2025 I think -- I get
-   KW: DOB \$44.65 all private in 2023

```{r}
#| label: setup
#| output: false

source(fs::path(DMAIN, "R", "taxrules_and_forecast_functions.R"))

# a <- str2expression("5 < 10")
# eval(a)

```

```{r}
#| label: get-data
#| output: false

pmtdb <- readRDS(fs::path(PDINTERMEDIATE, "pmtdb_calibrated.rds"))
glimpse(pmtdb)
summary(pmtdb)

```

Quick checks

```{r}

df <- readRDS(fs::path(PDINTERMEDIATE, "qcewmta.rds"))
count(df, area)

df |> 
  filter(ownerf==1, naics=="00") |> 
  summarise(payroll = sum(payroll)) |> 
  mutate(tax1bp=.0001 * payroll)

```

```{r}
#| label: problem-setup
#| output: false

# prob$emp_growth <- 0
# prob$avg_wagerate_growth <- 0

```

### Tax at 2024 levels

```{r}
#| label: tax-calc1
#| output: false

plan2024 <- make_taxplan("tax2024")

prob <- list()
prob$year <- 2024
prob$emp_growth <- .0
prob$avg_wagerate_growth <- .0
prob$plan1 <- plan2024
prob$plan2 <- plan2024

prob

pmtfc <- pmtdb_forecast(prob, pmtdb)
summary(pmtfc)

check <- pmtfc |> 
    mutate(calc_tax(pick(everything()), prob$plan1))

# sum(check$tax) # 2,945,979,702 # at 2024 levels
check |> 
  summarise(tax=sum(tax), .by=c(year, owner, ownerf)) |> 
  adorn_totals()


check |> 
  filter(gpay_firm > 10e6) |> 
  summarise(tax=sum(tax), .by=c(year, owner, ownerf)) |> 
  adorn_totals()

 #              owner ownerf        tax
 #            Private      1 2694159429
 # Federal Government      2          0
 #   State Government      3   46345527
 #   Local Government      4  205474746
 #              Total     10 2945979702

# wage tax target from calibration
# wagetax2024_target <- 2.946 * 1e9 # pulled from data.ny.gov 3/31/2025
#  2,946,047 
# my uncalibrated 2024 number is $3,172, which is $226 million or 7.7% higher
```

### 1 bp Tax at 2024 levels

```{r}
#| label: tax-calc2
#| output: false

plan1bp <- make_taxplan("tax1bp",
             lower_nyc = c(0, 1.5e6, 1.75e6, 10e6),
             rates_nyc = c(0.0001, 0.0001, 0.0001, 0.0001),
             lower_burbs = c(0, 1.5e6, 1.75e6, 10e6),
             rates_burbs = c(0.0001, 0.0001, 0.0001, 0.0001),
             description = "Tax at 1 bp")

prob <- list()
prob$year <- 2024
prob$emp_growth <- .0
prob$avg_wagerate_growth <- .0
prob$plan1 <- plan1bp
prob$plan2 <- plan1bp

prob

pmtfc <- pmtdb_forecast(prob, pmtdb)
summary(pmtfc)

check <- pmtfc |> 
    mutate(calc_tax(pick(everything()), prob$plan1))

check |> 
  summarise(tax=sum(tax), .by=c(owner, ownerf)) |> 
  adorn_totals()

check |> 
  filter(gpay_firm > 10e6) |> 
  summarise(tax=sum(tax), .by=c(owner, ownerf)) |> 
  adorn_totals()

 #              owner ownerf      tax
 #            Private      1 49717666
 # Federal Government      2        0
 #   State Government      3  1001862
 #   Local Government      4  3760685
 #              Total     10 54480213


```

### Tax at 2023 levels

```{r}
#| label: tax-calc3
#| output: false

plan2024 <- make_taxplan("tax2024")

prob <- list()
prob$year <- 2024
prob$emp_growth <- -.018
prob$avg_wagerate_growth <- -.038
prob$plan1 <- plan2024
prob$plan2 <- plan2024

prob

pmtfc <- pmtdb_forecast(prob, pmtdb)
summary(pmtfc)

check <- pmtfc |> 
    mutate(calc_tax(pick(everything()), prob$plan1))

# sum(check$tax) # 2,945,979,702 # at 2024 levels
check |> 
  summarise(tax=sum(tax), .by=c(year, owner, ownerf)) |> 
  adorn_totals()

 #  year              owner ownerf        tax
 #  2024            Private      1 2537015158
 #  2024 Federal Government      2          0
 #  2024   State Government      3   43663357
 #  2024   Local Government      4  194108705
 # Total                  -     10 2774787221


check |> 
  filter(gpay_firm > 10e6) |> 
  summarise(tax=sum(tax), .by=c(year, owner, ownerf)) |> 
  adorn_totals()

 #  year              owner ownerf        tax
 #  2024            Private      1 2077135933
 #  2024 Federal Government      2          0
 #  2024   State Government      3   27283543
 #  2024   Local Government      4  111344718
 # Total                  -     10 2215764193

# wage tax target from calibration
# wagetax2024_target <- 2.946 * 1e9 # pulled from data.ny.gov 3/31/2025
#  2,946,047 
# my uncalibrated 2024 number is $3,172, which is $226 million or 7.7% higher
```

### 1 bp Tax at 2023 levels

```{r}
#| label: tax-calc4
#| output: false

# back out my 2023 to 2024 assumptions:
#   Employment growth: 1.8%
# Growth in wages per employees: 3.8%
# gpayroll * (1 + prob$avg_wagerate_growth) * (1 + prob$emp_growth)) 

plan1bp <- make_taxplan("tax1bp",
             lower_nyc = c(0, 1.5e6, 1.75e6, 10e6),
             rates_nyc = c(0.0001, 0.0001, 0.0001, 0.0001),
             lower_burbs = c(0, 1.5e6, 1.75e6, 10e6),
             rates_burbs = c(0.0001, 0.0001, 0.0001, 0.0001),
             description = "Tax at 1 bp")

prob <- list()
prob$year <- 2023
prob$emp_growth <- -.018
prob$avg_wagerate_growth <- -.038
prob$plan1 <- plan1bp
prob$plan2 <- plan1bp

prob

pmtfc <- pmtdb_forecast(prob, pmtdb)
summary(pmtfc)

check <- pmtfc |> 
    mutate(calc_tax(pick(everything()), prob$plan1))

# 1 bp at 2023 levels
check |> 
  summarise(tax=sum(tax), .by=c(owner, ownerf)) |> 
  adorn_totals()

 #              owner ownerf      tax
 #            Private      1 49717666
 # Federal Government      2        0
 #   State Government      3  1001862
 #   Local Government      4  3760685
 #              Total     10 54480213

# 1 bp at 2023 levels

check |> 
  filter(gpay_firm > 10e6) |> 
  summarise(tax=sum(tax), .by=c(owner, ownerf)) |> 
  adorn_totals()

 #              owner ownerf        tax
 #            Private      1 37122512.5
 # Federal Government      2        0.0
 #   State Government      3   547776.7
 #   Local Government      4  2038493.1
 #              Total     10 39708782.3

```

## DOB option

Changes private

```{r}

plan2024 <- make_taxplan("tax2024")

plandob <- make_taxplan("dob2025",
             lower_nyc = c(0, 1.5e6, 1.75e6, 10e6),
             rates_nyc = c(0.00055, 0.00115, 0.006, 0.00895),
             lower_burbs = c(0, 1.5e6, 1.75e6, 10e6),
             rates_burbs = c(0.00055, 0.00115, 0.0034, 0.00635),
             description = "DOB 2025-04-28 plan")



```

## NOT UPDATED BELOW HERE

```{r}
#| label: show-dob-results
#| output: true

tabdata <- taxcomp |> 
  summarise(gpayroll = sum(gpayroll),
            tax=sum(tax),
            .by=c(ownerf, owner, basealt)) |> 
  pivot_wider(names_from = basealt,
              values_from = tax) |> 
  select(ownerf, owner, gpayroll, basetax, alttax) |> 
  arrange(ownerf) |> 
  adorn_totals() |> 
  mutate(change = alttax - basetax)

tabdata |> 
  gt() |> 
  tab_header(md("50bp PMT increase for **firms** with payroll > $10m, NYC+Suburbs combined"),
             subtitle="$ millions, 2025 levels") |> 
  fmt_number(columns = -c(ownerf, owner),
             scale = 1e-6,
             decimals = 0) |> 
  fmt_currency(columns = -c(ownerf, owner),
               rows = c(1, nrow(tabdata)),
               scale = 1e-6,
               decimals = 0)

```

### Results broken down by NYC and suburbs

```{r}
#| label: show-dob-geographic-details
#| output: true

tabdata <- taxcomp |> 
  summarise(tax=sum(tax),
            .by=c(nyc, ownerf, owner, basealt)) |> 
  mutate(nyc=ifelse(nyc, "nyc", "suburbs")) |> 
  pivot_wider(names_from = basealt,
              values_from = tax) |> 
  mutate(change = alttax - basetax) |> 
  pivot_wider(names_from = nyc,
              values_from = c(basetax, alttax, change)) |> 
  arrange(ownerf) |> 
  adorn_totals() |> 
  mutate(base_total = basetax_nyc + basetax_suburbs,
         alt_total = alttax_nyc + alttax_suburbs,
         change_total = change_nyc + change_suburbs,
         nyc_change_pct=change_nyc / change_total) |> 
  relocate(base_total, .after = basetax_suburbs) |> 
  relocate(alt_total, .after = alttax_suburbs)
# tabdata

tabdata |> 
  gt() |> 
  tab_header(md("50bp PMT increase for **firms** with payroll > $10m"),
             subtitle="$ millions, 2025 levels") |> 
  fmt_number(columns = -c(ownerf, owner,nyc_change_pct),
             scale = 1e-6,
             decimals = 0) |> 
  fmt_currency(columns = -c(ownerf, owner,nyc_change_pct),
               rows = c(1, nrow(tabdata)),
               scale = 1e-6,
               decimals = 0) |> 
  fmt_percent(columns=nyc_change_pct,
              decimals = 1) |> 
  sub_missing(missing_text = "--")

```

### Plan 1 (2024 law) and Plan 2 (DOB option) details

```{r}
#| label: show-plans
#| output: true

prob

```

## Save data for MTA

```{r}
#| label: payroll-cutpoints
#| output: false

# define payroll cuts and labels
legal_cuts <- 4 * c(312.5e3, 375e3, 437.5e3)

(payroll_cuts <- c(0, 500e3, seq(1e6, 2e6, 200e3), seq(2e6, 10e6, 500e3), Inf, legal_cuts) |> unique() |> sort())
paygroup_labels <- cutlabs(payroll_cuts)

```

```{r}
#| label: prepare-mta-data
#| output: false

glimpse(taxcomp)
count(taxcomp, shortname, basealt)

for_mta_prep <- taxcomp |> 
  left_join(paygroup_labels |> 
              select(paygroup, lb, ub, paylabel = range_left), 
            # rstudio shows syntax error on between -- it does not like bounds -- but it's valid
            by = join_by(between(gpay_firm, lb, ub, bounds = "[)"))) |> 
  select(-c(lb, ub)) |> 
  arrange(year, naics, exempt, excluded, paygroup, paylabel, desc(nyc), area, desc(basealt)) 

# check <- for_mta_prep |> 
#   filter(area=="Bronx", ownerf==1, paygroup==1, !low_pay, naics==236)


for_mta <- for_mta_prep |> 
  mutate(area=ifelse(nyc, "NYC", area)) |> 
  summarise(across(c(gfirms, gestabs, gemp, gpayroll, tax), sum),
            .by=c(basealt, year, excluded, exempt, low_pay, nyc, area, ownerf, owner, paygroup, paylabel, rate)) |>
  pivot_wider(names_from = basealt, values_from = c(rate, tax)) |> 
  mutate(taxchange = tax_alttax - tax_basetax) |> 
  arrange(year, paygroup, paylabel, excluded, exempt, low_pay, desc(nyc), area)

summary(for_mta)
sum(for_mta$taxchange)

```

```{r}
#| label: write-mta-data
#| output: false

library(openxlsx2)

template_fname <- "template_database_summary.xlsx"
template_path <- fs::path(DMAIN, "data_for_mta", template_fname)

wb <- openxlsx2::wb_load(template_path, sheet = "database")

firstrow <- 6

wb <- openxlsx2::wb_add_data(
  wb = wb,
  sheet = "database",
  x = for_mta,
  start_row = firstrow,
  col_names = FALSE,
  apply_cell_style = FALSE
)

fmt_dims <- wb_dims(cols = c("K", "L", "M", "N", "Q", "R", "S"), rows = firstrow:(firstrow + nrow(for_mta) - 1))
wb <- wb_add_numfmt(
  wb,
  sheet = "database",
  dims = fmt_dims,
  numfmt = "#,##0"
)

fmt_dims <- wb_dims(cols = c("O", "P"), rows = firstrow:(firstrow + nrow(for_mta) - 1))
wb <- wb_add_numfmt(
  wb,
  sheet = "database",
  dims = fmt_dims,
  numfmt = "0.000%"
)

fnbase <- "pmt_summary_database_with_base_and_dob_tax"
fname <- paste0(fnbase, "_", Sys.Date(), ".xlsx")
fpath <- fs::path(DMAIN, "data_for_mta", fname)
openxlsx2::wb_save(wb, file = fpath, overwrite = TRUE)


```

## Show summary of data file prepared for MTA

```{r}
#| label: show-mta-data-summary
#| output: true

for_mta |> 
  summarise(across(c(gpayroll, tax_basetax, tax_alttax, taxchange), sum),
            .by=c(ownerf, owner)) |> 
  arrange(ownerf) |> 
  adorn_totals() |>  
  gt() |> 
  tab_header("Payroll and tax amounts, 2024 law and DOB proposal for 50bp increase firms with >= $10m payroll",
             subtitle = "MTA region") |> 
  fmt_number(columns = gpayroll:taxchange, decimals = 0)

```

## Example 3: Increase rate by 36 bps in entire region for all PRIVATE and PUBLIC employers greater than \$10 million, exempt all with payroll less than \$2 MM

-   Raise top PMT rate by 50 basis points on firms with payroll \>= \$10 million
-   No change for firms with lower payroll

```{r}
#| label: problem-setup-ex3
#| output: false

plan2024 <- make_taxplan("tax2024")

ex3 <- make_taxplan("ex3", 
             lower_nyc = c(0, 2e6, 10e6),
             rates_nyc = c(0.0, 0.006, .006 + .0036),
             lower_burbs = c(0, 2e6, 10e6),
             rates_burbs = c(0.0, 0.0034, 0.0034 + .0036),
             description = "Increase rate by 36 bps in entire region >= $10m, exempt < 2m")

# plandob <- make_taxplan("dob10m", 
#              lower_nyc = c(0, 1.5e6, 1.75e6, 10e6),
#              rates_nyc = c(0.0011, 0.0023, 0.006, 0.011),
#              lower_burbs = c(0, 1.5e6, 1.75e6, 10e6),
#              rates_burbs = c(0.0011, 0.0023, 0.0034, 0.0084),
#              description = "DOB rate increase for payroll >= $10m")


prob <- list()
prob$year <- 2025
prob$emp_growth <- .015
prob$avg_wagerate_growth <- .04
prob$plan1 <- plan2024
prob$plan2 <- ex3

prob

# prob$emp_growth <- 0
# prob$avg_wagerate_growth <- 0

```

### Summary results

```{r}
#| label: tax-calcx3
#| output: false

pmtfc <- pmtdb_forecast(prob, pmtdb)
summary(pmtfc)

taxcomp <- bind_rows(
  pmtfc |> 
    mutate(calc_tax(pick(everything()), prob$plan1)) |> 
    mutate(basealt = "basetax"),
  pmtfc |> 
    mutate(calc_tax(pick(everything()), prob$plan2)) |> 
    mutate(basealt = "alttax"))

glimpse(taxcomp)
count(taxcomp, shortname)

taxcomp |> 
  summarise(tax=sum(tax),
            .by=c(ownerf, owner, shortname)) |> 
  pivot_wider(names_from = shortname,
              values_from = tax) |> 
  adorn_totals() |> 
  mutate(diff = ex3 - tax2024)

```

```{r}
#| label: show-ex3-results
#| output: true

tabdata <- taxcomp |> 
  summarise(gpayroll = sum(gpayroll),
            tax=sum(tax),
            .by=c(ownerf, owner, basealt)) |> 
  pivot_wider(names_from = basealt,
              values_from = tax) |> 
  select(ownerf, owner, gpayroll, basetax, alttax) |> 
  arrange(ownerf) |> 
  adorn_totals() |> 
  mutate(change = alttax - basetax)

tabdata |> 
  gt() |> 
  tab_header(md("Increase rate by 36 bps in entire region >= $10m, exempt < 2m"),
             subtitle="$ millions, 2025 levels") |> 
  fmt_number(columns = -c(ownerf, owner),
             scale = 1e-6,
             decimals = 0) |> 
  fmt_currency(columns = -c(ownerf, owner),
               rows = c(1, nrow(tabdata)),
               scale = 1e-6,
               decimals = 0)

```

### Results broken down by NYC and suburbs

```{r}
#| label: show-dob-geographic-details
#| output: true

tabdata <- taxcomp |> 
  summarise(tax=sum(tax),
            .by=c(nyc, ownerf, owner, basealt)) |> 
  mutate(nyc=ifelse(nyc, "nyc", "suburbs")) |> 
  pivot_wider(names_from = basealt,
              values_from = tax) |> 
  mutate(change = alttax - basetax) |> 
  pivot_wider(names_from = nyc,
              values_from = c(basetax, alttax, change)) |> 
  arrange(ownerf) |> 
  adorn_totals() |> 
  mutate(base_total = basetax_nyc + basetax_suburbs,
         alt_total = alttax_nyc + alttax_suburbs,
         change_total = change_nyc + change_suburbs,
         nyc_change_pct=change_nyc / change_total) |> 
  relocate(base_total, .after = basetax_suburbs) |> 
  relocate(alt_total, .after = alttax_suburbs)
# tabdata

tabdata |> 
  gt() |> 
  tab_header(md("50bp PMT increase for **firms** with payroll > $10m"),
             subtitle="$ millions, 2025 levels") |> 
  fmt_number(columns = -c(ownerf, owner,nyc_change_pct),
             scale = 1e-6,
             decimals = 0) |> 
  fmt_currency(columns = -c(ownerf, owner,nyc_change_pct),
               rows = c(1, nrow(tabdata)),
               scale = 1e-6,
               decimals = 0) |> 
  fmt_percent(columns=nyc_change_pct,
              decimals = 1) |> 
  sub_missing(missing_text = "--")

```

```{r}
#| label: cleanup
rm(list = ls())
```

```{r stop_here, echo=FALSE}
knitr::knit_exit()
```
