# DOB PMT options

```{r}
#| label: setup

suppressPackageStartupMessages(source(here::here("R", "libraries.r")))
suppressPackageStartupMessages(source(here::here("R", "libraries_ts.r")))

source(here::here("R", "constants.r"))
source(here::here("R", "functions.r"))
source(here::here("pmt", "pmt_constants_and_functions.r"))

# GLOBAL OPTIONS:
#  eval: true
#  output: false
#  echo: false

```

```{r}
#| label: prepare-pmt-data
#| output: false

pmtbl <- readRDS(fs::path(PDINTERMEDIATE, "pmt2024_baseline.rds"))
glimpse(pmtbl)
count(pmtbl, status)

mta <- readRDS(fs::path(PDINTERMEDIATE, "susb2021_mta.rds"))

(payroll_cuts <- c(0, 500e3, seq(1e6, 2e6, 200e3), seq(2e6, 5e6, 1e6), 10e6, Inf, 1.25e6, 1.75e6) |> unique() |> sort())
paygroup_labels <- cutlabs(payroll_cuts)

```

## Option: Exclude businesses with payroll below \$10 million and raise rate by 50 bps in City or entire MTA

Proposal that would exempt PMT below \$10 million payroll and raise rate by 50 bps.

Purported estimate of \$2 billion if City only and \$2.4 billion if suburbs included.

### Analysis using PMT database

```{r}

#   payroll_ub    nyc suburbs
#        <dbl>  <dbl>   <dbl>
# 1    1250000 0       0     
# 2    1500000 0.0011  0.0011
# 3    1750000 0.0023  0.0023
# 4        Inf 0.006   0.0034

ten_million_treshold_plus50bps <- 
  tribble(
    ~payroll_ub, ~nyc, ~suburbs,
    10e6, 0, 0,
    Inf, 0.011, 0.0084
  )
  
ten_million_treshold_plus50bps |> knitr::kable()

pmtbl_alt <- pmtbl |> 
  mutate(paygroup=cut(esize_payestab, payroll_cuts, right=FALSE, labels=FALSE),
         paylabel=getlabel(paygroup, paygroup_labels)) |> 
  arrange(paygroup)


alt <- pmtbl_alt |> 
  mutate(get_info(esize_payestab, nyc, rates=ten_million_treshold_plus50bps, prefix="alt_"),
         alt_tax= esize_payroll * alt_rate)

alt |>
  summarise(n=n(),
          across(c(base_tax, alt_tax),
                 \(x) sum(x)),
                 .by=c(paygroup, paylabel)) |> 
  janitor::adorn_totals() |> 
  mutate(change=alt_tax - base_tax) |> 
  knitr::kable(decimals=1, format.args = list(big.mark = ","))


data1 <- alt |> 
  summarise(n=n(),
            across(c(esize_estabsadj, 
                     esize_emp,
                     base_tax,
                     alt_tax),
                   \(x) sum(x)),
                   .by=c(nyc, paygroup, paylabel))

tabdata <- data1 |> 
  select(-n, -esize_emp) |> 
  mutate(nyc=factor(nyc, levels=c(TRUE, FALSE), labels = c("NYC", "Suburbs"))) |> 
  pivot_wider(names_from=nyc, values_from = c(esize_estabsadj, base_tax, alt_tax)) |> 
  mutate(base_total_tax = base_tax_NYC + base_tax_Suburbs,
         alt_total_tax = alt_tax_NYC + alt_tax_Suburbs) |> 
  arrange(paygroup) |> 
  select(paylabel, contains("NYC"), contains("Suburbs"), base_total_tax, alt_total_tax) |> 
  janitor::adorn_totals() |> 
  mutate(change=alt_total_tax - base_total_tax,
         nyc_change_share = (alt_tax_NYC - base_tax_NYC) / change)

```

```{r}
#| label: show-10-million-option
#| output: true
#| column: page

tabdata |> 
  gt() |> 
  tab_header("$10 million threshold and 50 basis point increase in rate",
             subtitle = "Private sector only. Amounts in $ millions") |> 
  cols_label(paylabel="Average establishment payroll ($)",
             contains("estabsadj") ~ "# of establishments",
             contains("base_tax") ~ "Baseline PMT wage tax ($m)",
             contains("alt_tax") ~ "Alternative PMT wage tax ($m)",
             base_total_tax = "Baseline total PMT wage tax ($m)",
             alt_total_tax = "Alternative total tax ($m)",
             change = "Change in tax ($m)",
             nyc_change_share = "NYC share of tax change") |> 
  tab_spanner(label="New York City", columns=contains("NYC", ignore.case = FALSE)) |> 
  tab_spanner(label="Suburbs", columns=contains("suburbs", ignore.case = TRUE)) |> 
  fmt_number(columns = contains("estabsadj"), decimals=0) |>
  fmt_number(columns = c(contains("tax"), change), scale = 1e-6, decimals = 1) |>
  fmt_currency(rows=c(1, nrow(tabdata)), columns = c(contains("tax"), change), scale=1e-6, decimals=1) |> 
  fmt_percent(columns = c(nyc_change_share), decimals=1)  |> 
  sub_missing(missing_text = "--") |> 
  tab_source_note(source_note = source_note) |> 
  tab_source_note(source_note = long_note)

```

### SUSB sanity-checking the option

Get SUSB (Statistics for U.S. Businesses) for the MTA region.

This should give overestimate of wages.

```{r}
#| output: false

# saveRDS(susbdtl3, fs::path(PDINTERMEDIATE, "susb2021_usstate_details.rds"))

glimpse(mta)
count(mta, cofips, county)
count(mta, fscode, fslb, fsub)
count(mta, firmsize)
tmp <- mta |> count(naics, description)

mta2 <- mta |> 
  mutate(lnaics = ifelse(naics == "--", 1, nchar(naics)),
         fsgroup = ifelse(fscode==1, 1, 2),
         ppf=payroll / firms)

```

#### SUSB payroll checks, different industry summary levels

```{r}
#| output: true

mta2 |> 
  summarise(firms=sum(firms), 
            payroll=sum(payroll),
            .by=c(fsgroup, lnaics))

```

#### Payroll calibration for SUSB to match PMT database

```{r}

# sum(pmtbl$esize_payroll) 627744347597
qcewpayroll <- 627744347597 # 627.744 billion
susbpayroll <- 505127862000
# here's what I get from susb spreadsheet 505,127,862 -- good
adjust <- qcewpayroll / susbpayroll

# $1,759 tax-increase estimate grows to $2,205.6 million if we adjust upward by adjust 1.242743

```

#### SUSB payroll and tax 50bp-plus checks, different industry summary levels and firmsize groups

```{r}
#| output: true

mta2 |> 
  mutate(ppf = ppf * adjust,
         payroll = payroll * adjust) |>
  filter(ppf >= 10e6) |> 
  mutate(tax50=.0050 * payroll) |> 
  summarise(firms=sum(firms), 
            payroll=sum(payroll),
            tax50=sum(tax50),
            .by=c(lnaics, fsgroup)) |> 
  arrange(lnaics, fsgroup)

```

#### SUSB quick and dirty model

```{r}

susb_base <- mta2 |>  
  mutate(ppf = ppf * adjust,
         payroll = payroll * adjust) |> 
  filter(lnaics==1, fsgroup==2) |> 
  mutate(baserate = case_when(
    nyc & ppf <= 1.25e6 ~ 0,
    nyc & ppf <= 1.5e6 ~ 0.0011,
    nyc & ppf <= 1.75e6 ~ 0.0023,
    nyc & ppf > 1.75e6 ~ 0.006,
    # suburbs
    !nyc & ppf <= 1.25e6 ~ 0,
    !nyc & ppf <= 1.5e6 ~ 0.0011,
    !nyc & ppf <= 1.75e6 ~ 0.0023,
    !nyc & ppf > 1.75e6 ~ 0.0034,
    .default = 1e-9),
    basetax = payroll * baserate)


```

Base tax rates and base tax

```{r}
#| output: true

base_rates

susb_base |> 
  summarise(payroll=sum(payroll),
            basetax=sum(basetax))

```

Alternative tax

-   Exclude firms from tax with \< \$10 million payroll
-   50bp rate increase for firms with \>= \$10m payroll

```{r}
#| output: true
  
susb_alt <- susb_base |> 
  mutate(altrate = case_when(
    ppf <= 10e6 ~ 0,
    ppf > 10e6 ~ baserate + 0.005,
    .default = 1e-9),
    alttax = payroll * altrate)

susb_alt |> 
  mutate(diff = alttax - basetax,
         change= ifelse(diff > 0, "up", "down")) |> 
  summarise(payroll=sum(payroll),
            basetax=sum(basetax),
            alttax = sum(alttax),
            diff = sum(diff),
            .by=change) |> 
  janitor::adorn_totals()

```
