---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Get qcew data

```{r}
#| label: setup
#| include: false

source(here::here("R", "libraries.r"))
source(here::here("R", "libraries_ts.r"))
source(here::here("R", "constants.r"))
source(here::here("R", "functions.r"))

library(zip)
library(lobstr)

```


## QCEW data from the New York State Department of Labor

The New York State Department of Labor (NYSDOL) provides a version of the BLS Quarterly Census of Employment and Wages (QCEW) that has quarterly county-level wage, establishment, and employment data in a format we need. The data are available from 2000 through a recent quarter. A new quarter and revisions to prior quarters generally are available about 6 months after end of a quarter.

```{r}
#| label: qcew-names

url <- "https://statistics.labor.ny.gov/qcew.zip"
fname <- fs::path_file(url)
zpath <- fs::path(PMTDATARAW, fname)
zpath <- paste0(PMTDATARAW, "/", fname)

```

### Download qcew data

Note: The latest qcew.zip file, which I downloaded once manually and once with the code below, appears to have been corrupted, at least as downloaded by my machine. I had to use an external utility (WinRar) to unzip it to a separate folder, then rezip it, after which the computer could read it.

```{r}
#| label: download-qcew
#| eval: false

# Sys.setenv(R_DEFAULT_INTERNET_TIMEOUT = 400)  
options(timeout = max(400, getOption("timeout")))
download.file(url, zpath, mode="wb")

```

### Read and save QCEW data

```{r}
#| label: read-save-qcew
#| output: false

zfiles <- zip::zip_list(zpath)

csvfiles <- zfiles |> 
  filter(str_starts(filename, "qcew_annual_")) |> 
  pull(filename)

# read directly from the zip file
system.time(df <- vroom(purrr::map(csvfiles, \(x)  unz(zpath, x)))) # 2.8 secs
# obj_size(df)

df2 <- df |> 
  rename_with(str_to_lower)

# count(df2, areatype)
# count(df2, areatype, owner)

df3 <- df2 |> 
  filter(areatype %in% c("State", "County")) |> 
  mutate(areatype=str_to_lower(areatype)) |> 
  mutate(mta=area %in% constants$mtaregion$area,
         nyc=area %in% constants$nyc$area)

saveRDS(df3, fs::path(PMTDINTERMEDIATE, "qcewstco"))

rm(df, df2, df3)

```


### Explore qcew

```{r}
#| label: explore

qcew <- readRDS(fs::path(PMTDINTERMEDIATE, "qcewstco"))

count(qcew |> filter(areatype=="state"), area)
count(qcew |> filter(areatype=="county"), area)
# 65: 57 + 5 NYC +
#     Locations Statewide (16) + Out of State Locations (2) + Unknown Locations (16)

count(qcew, year)
count(qcew, owner)
check <- count(qcew, naics, naics_level, title)

# check sums
qcew |> 
  filter(year==2023, naics=="00", owner=="Total Private and Government") |> 
  summarise(totwage=sum(totwage),
            avgemp=sum(avgemp), 
            estab=sum(estab),
            .by=c(areatype)) |> 
  pivot_longer(-areatype) |> 
  pivot_wider(names_from = areatype) |> 
  mutate(pct=county / state)

#   name           state       county   pct
#   <chr>          <dbl>        <dbl> <dbl>
# 1 totwage 866004485600 844079889657 0.975
# 2 avgemp       9472105      9293900 0.981
# 3 estab         697828       640188 0.917

# count(df2, areatype)
#   areatype                            n
#   <chr>                           <int>
# 1 Balance of State               101498
# 2 County                        2066833
# 3 Labor Market Area              741876
# 4 Metropolitan Division          140696
# 5 Metropolitan Statistical Area  716101
# 6 Micropolitan Statistical Area  292755
# 7 State                          112968
# 8 Workforce Investment Region    103316
  

# count(df2, areatype, owner)
# 37 State                         Federal Government              2166
# 38 State                         Local Government                2743
# 39 State                         Private                        48547
# 40 State                         State Government                1664
# 41 State                         Total Government                6658
# 42 State                         Total Private and Government   51190

```

