---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Get economic data

```{r}
#| label: setup
#| include: false

source(here::here("R", "libraries.r"))
source(here::here("R", "libraries_ts.r"))
source(here::here("R", "constants.r"))
source(here::here("R", "functions.r"))

library(zip)
library(lobstr)

```


```{r}
#| label: constants

PMTDATA <- here::here("pmt", "data")
PMTDATARAW <- fs::path(PMTDATA, "data_raw")
PMTDINTERMEDIATE <- fs::path(PMTDATA, "intermediate")

```


## QCEW data from the New York State Department of Labor

The New York State Department of Labor (NYSDOL) provides a version of the BLS Quarterly Census of Employment and Wages (QCEW) that has quarterly county-level wage, establishment, and employment data in a format we need. The data are available from 2000 through a recent quarter. A new quarter and revisions to prior quarters generally are available about 6 months after end of a quarter.

```{r}
#| label: qcew-names

url <- "https://statistics.labor.ny.gov/qcew.zip"
fname <- fs::path_file(url)
zpath <- fs::path(PMTDATARAW, fname)
zpath <- paste0(PMTDATARAW, "/", fname)

```

### Download qcew data

Note: The latest qcew.zip file, which I downloaded once manually and once with the code below, appears to have been corrupted, at least as downloaded by my machine. I had to use an external utility (WinRar) to unzip it to a separate folder, then rezip it, after which the computer could read it.

```{r}
#| label: download-qcew
#| eval: false

# Sys.setenv(R_DEFAULT_INTERNET_TIMEOUT = 400)  
options(timeout = max(400, getOption("timeout")))
download.file(url, zpath, mode="wb")

```

### Read and save QCEW data

```{r}
#| label: read-save-qcew
#| output: false

zfiles <- zip::zip_list(zpath)

csvfiles <- zfiles |> 
  filter(str_starts(filename, "qcew_annual_")) |> 
  pull(filename)

# read directly from the zip file
system.time(df <- vroom(purrr::map(csvfiles, \(x)  unz(zpath, x)))) # 2.8 secs
# obj_size(df)

df2 <- df |> 
  rename_with(str_to_lower)

# count(df2, areatype)
# count(df2, areatype, owner)

df3 <- df2 |> 
  filter(areatype %in% c("State", "County")) |> 
  mutate(areatype=str_to_lower(areatype)) |> 
  mutate(mta=area %in% constants$mtaregion$area,
         nyc=area %in% constants$nyc$area)

saveRDS(df3, fs::path(PMTDINTERMEDIATE, "qcewstco"))

rm(df, df2, df3)

```


### Explore qcew

```{r}

qcew <- readRDS(fs::path(PMTDINTERMEDIATE, "qcewstco"))

count(qcew |> filter(areatype=="state"), area)
count(qcew |> filter(areatype=="county"), area)
# 65: 57 + 5 NYC +
#     Locations Statewide (16) + Out of State Locations (2) + Unknown Locations (16)

count(qcew, year)
count(qcew, owner)
check <- count(qcew, naics, naics_level, title)

# check sums
qcew |> 
  filter(year==2023, naics=="00", owner=="Total Private and Government") |> 
  summarise(totwage=sum(totwage),
            avgemp=sum(avgemp), 
            estab=sum(estab),
            .by=c(areatype)) |> 
  pivot_longer(-areatype) |> 
  pivot_wider(names_from = areatype) |> 
  mutate(pct=county / state)

#   name           state       county   pct
#   <chr>          <dbl>        <dbl> <dbl>
# 1 totwage 866004485600 844079889657 0.975
# 2 avgemp       9472105      9293900 0.981
# 3 estab         697828       640188 0.917

# count(df2, areatype)
#   areatype                            n
#   <chr>                           <int>
# 1 Balance of State               101498
# 2 County                        2066833
# 3 Labor Market Area              741876
# 4 Metropolitan Division          140696
# 5 Metropolitan Statistical Area  716101
# 6 Micropolitan Statistical Area  292755
# 7 State                          112968
# 8 Workforce Investment Region    103316
  

# count(df2, areatype, owner)
# 37 State                         Federal Government              2166
# 38 State                         Local Government                2743
# 39 State                         Private                        48547
# 40 State                         State Government                1664
# 41 State                         Total Government                6658
# 42 State                         Total Private and Government   51190

```


## TODO: Make PMT tax base data

```{r}



```


## OLD QCEW STUFF

```{r}
#| label: ONETIME-get-save-nysdol-qcew
#| eval: false
#| include: false

qurl <- "https://statistics.labor.ny.gov/qcew.zip"
fname <- fs::path_file(qurl)
download.file(qurl, destfile = here::here("data", "nyqcew", fname), mode="wb")

```


```{r}
#| label: OLD-ONETIME-get-save-nysdol-qcew
#| eval: false
#| include: false

# caution: NYSDOL uses Unsupported ZIP compression method (9: deflation-64-bit)
# either convert to regular zip using winrar (or some other method)
# or extract and read

# qver <- "qcew_dl2023-05-28.zip"
qver <- "qcew_dl2023-10-12.zip"
qver <- "qcew_dl2024-06-11.zip"
# qver <- "E:\R_projects\MTA\MTA_forecast\forecast_2024-06\data"

qurl <- "https://statistics.labor.ny.gov/qcew.zip"
fname <- fs::path_file(qurl)
download.file(qurl, destfile = here::here("data", "nyqcew", fname), mode="wb")

pzip <- here::here("data", "nyqcew", qver)

# files <- fs::dir_ls(path=here::here("data", "nyqcew"), glob = "*csv")
(files <- unzip(pzip, list = TRUE)$Name |>
    str_subset(".csv") |> 
    str_subset("quarter"))

# files <- fs::dir_ls(path=here::here("data", "nyqcew"), glob = "*csv")

files2 <- path(here::here("data", ""))

df <- vroom(purrr::map(files[1], ~ unz(pzip, .x)))

vroom(pzip:files)
vroom(here::here("data", "nycqew", files[1]))

# E:\R_projects\MTA\MTA_forecast\data\nyqcew
# E:\R_projects\MTA\MTA_forecast\data\nyqcew
# E:/R_projects/MTA/MTA_forecast/data/nycqew
fp <- fs::path("E:/R_projects/MTA/MTA_forecast/data/nycqew", files[1])
fp <- r"(E:\R_projects\MTA\MTA_forecast\data\nyqcew\qcew_quarter_2000.csv)"
(paths <- paste0("E:\\R_projects\\MTA\\MTA_forecast\\data\\nyqcew\\", files))
df <- vroom(paths)
glimpse(df)
count(df, Owner)

df2 <- df |> 
  filter(NAICS=="00", Owner=="Total Private and Government") |> 
  lcnames() |> 
  select(areatype, area, year, quarter, totwage) |> 
  mutate(date=yq(paste(year, quarter, sep = "-")))
glimpse(df2)

count(df2, date)
count(df2, areatype, area)

areas <- c("Bronx County",
           "Kings County",
           "New York County",
           "Queens County",
           "Richmond County",
           
           "Dutchess County",
           "Nassau County",
           "Orange County",
           "Putnam County",
           "Rockland County",
           "Suffolk County",
           "Westchester County"
           )

df3 <- df2 |> 
  filter(areatype=="County", area %in% areas) |> 
  mutate(group=ifelse(area %in% areas[1:5],
                      "nyc",
                      "suburbs"))

count(df3, group, area)

df4 <- df3 |> 
  summarise(n=n(),
            totwage=sum(totwage),
            .by=c(group, date))
tots <- df4 |> 
  summarise(n=n(),
            totwage=sum(totwage),
            .by=c(date)) |> 
  mutate(group="mta")

df5 <- bind_rows(df4, tots) |> 
  arrange(group, date) |> 
  mutate(pchya=totwage / lag(totwage, 4) - 1, .by=group)
df5


p <- df5 |> 
  filter(year(date)>=2015) |> 
  ggplot(aes(date, pchya, colour=group)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_y_continuous(name=NULL, breaks=seq(-.3, .3, .02), labels = scales::percent_format(accuracy = 1)) +
  scale_x_date(name=NULL, date_breaks = "1 year", labels = scales::date_format(format="%Y")) +
  ggtitle("MTA region year-over-year growth in non-seasonally adjusted wages",
          subtitle = "Source: QCEW") +
  theme_bw()
p
ggsave(path(respath, "nswages.png"), plot=p)  

df5 |> 
  filter(year(date)>=2021) |> 
  select(group, date, pchya) |> 
  pivot_wider(names_from = group, values_from = pchya) |> 
  gt() |> 
  fmt_percent(columns=-date, decimals=1) |> 
  fmt_date(
    columns = date,
    date_style = "year_quarter"
  )


fpath <- here::here("data", "boyd_analysis_for_July2024_finplan.xlsx")
fpath <- r"(E:\R_projects\MTA\MTA_forecast\forecast_2024-06\data\boyd_analysis_for_July2024_finplan.xlsx)"
ihs1 <- read_excel(fpath, sheet="econ", range="a4:n128")
glimpse(ihs1)

ihs2 <- ihs1 |> 
  select(qdate=1, ihs=Total) |> 
  mutate(date=yq(qdate)) |> 
  select(date, ihs)

ihs3 <- inner_join(
  tots |> 
    select(date, qcew=totwage) |> 
    mutate(qcew=qcew / 1e6),
  ihs2, 
  by = join_by(date)) |> 
  pivot_longer(cols=c(qcew, ihs), names_to = "group") |> 
  arrange(group, date) |> 
  mutate(pchya=value / lag(value, 4) - 1, .by=group) |> 
  mutate(fgroup=factor(group, levels=c("ihs", "qcew"), labels=c("seas (IHS)", "notseas (QCEW)")))

p <- ihs3 |> 
  filter(year(date)>=2015) |> 
  ggplot(aes(date, pchya, colour=fgroup)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_y_continuous(breaks=seq(-.3, .3, .01), labels = scales::percent_format(accuracy = .1)) +
  ggtitle("MTA region year-over-year growth in seasonally adjusted and non-adjusted wages") +
  theme_bw()
p  
ggsave(path(respath, "ns_vs_sa_wages.png"), plot=p)  

ihs3 |> 
  filter(year(date)>=2021) |> 
  select(group, date, pchya) |> 
  pivot_wider(names_from = group, values_from = pchya) |> 
  gt() |> 
  fmt_percent(columns=-date, decimals=1) |> 
  fmt_date(
    columns = date,
    date_style = "year_quarter"
  )

library(archive)
archive_extract(pzip) # Unsupported ZIP compression method (9: deflation-64-bit)



unzip(pzip)

temp_dir <- tempdir()
utils::unzip(pzip, files=files, exdir = temp_dir)

dir.create(tmp <- tempfile())
list.dirs(tmp)
utils::unzip(pzip, files=files, exdir = tmp)


zip::unzip(pzip, files = files, overwrite = TRUE, junkpaths = FALSE, exdir = temp_dir)



vroom(pzip)


glimpse(df)
names(df)
# [1] "AREATYPE"    "AREA"        "YEAR"        "QUARTER"     "NAICS"       "NAICS_LEVEL" "NAICS_TITLE" "ESTAB"       "MNTH1EMP"    "MNTH2EMP"   
# [11] "MNTH3EMP"    "TOTWAGE"     "Owner"  
count(df, AREATYPE) # County, State
count(df |> filter(AREATYPE=="County"), AREA)
count(df, Owner)
# Owner                              n
# <chr>                          <int>
# 1 Federal Government            102215
# 2 Local Government              245916
# 3 Private                      7052970
# 4 State Government              134571
# 5 Total Government              558464
# 6 Total Private and Government 8068156


df1 <- df |> 
  rename_with(tolower) |> 
  filter(areatype %in% c("County", "State"),
         owner %in% c("Total Private and Government", "Local Government", "Federal Government")) |> 
  filter(naics=="00" | (naics_level==4 & naics=="6111"))
skim(df1)

saveRDS(df1, here::here("data", "nyqcew", "mtaqcew_raw.rds"))

```

```{r}
#| label: ONETIME-save-nysdol-qcew-slim
#| eval: false
#| include: false


mtawage1 <- readRDS(here::here("data", "nyqcew", "mtaqcew_raw.rds"))
count(mtawage1, year, quarter) |> ht()

mtawage2 <- mtawage1 |> 
  filter(area=="New York State" | area %in% constants$mtaregion$area) |> 
  mutate(date=yq(paste0(year, "-", quarter)),
         areatype=str_to_lower(areatype),
         region=case_when(area %in% constants$mtaregion$area[1:5] ~ "nyc",
                          area %in% constants$mtaregion$area[6:12] ~ "suburban",
                          area == "New York State" ~ "nys",
                          TRUE ~ "ERROR"),
         owncode=factor(owner, 
                        levels=c("Total Private and Government", "Local Government", "Federal Government"),
                        labels=c("total", "locgov", "fedgov")),
         ind=case_when(naics=="00" ~ "allind",
                       naics=="6111" ~ "k12"))
count(mtawage2, areatype, region, area)
count(mtawage2, owncode, owner)
count(mtawage2, ind, naics, naics_level, naics_title)

mtawage3 <- mtawage2 |> 
  select(date, year, qtr=quarter, region, area, ind, owncode, estab:totwage) |> 
  pivot_longer(cols=estab:totwage)

saveRDS(mtawage3, here::here("data", "nyqcew", "mtaqcew_slim.rds"))

```

```{r}
#| label: ONETIME-mta-wagebase
#| eval: false
#| include: false

mtawage3 <- readRDS(here::here("data", "nyqcew", "mtaqcew_slim.rds"))

mtawage3 |> 
  filter(area=="Dutchess County",
         ind=="allind",
         name=="totwage") |> 
  filter(year(date)>=2010) |> 
  ggplot(aes(date, value)) +
  geom_line() +
  geom_point() +
  geom_smooth() +
  scale_x_date(breaks = "3 months") +
  facet_wrap(~owncode, scales="free_y", ncol=1) +
  theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust=0.5)) 

check <- mtawage3  |> filter(area=="New York County", date==max(date))
check <- mtawage3  |> filter(region=="nyc", date==max(date))
count(check, area)
count(check, ind)
count(check, owncode)

xk12 <-  mtawage3 |> 
  filter(ind=="k12" & owncode=="locgov")

xk12all <-  mtawage3 |> 
  filter(ind=="k12")
xk12all |> filter(region=="nys", date==max(date))
xk12all |> filter(region=="nyc", date==max(date)) # no local gov k12 data for nyc -- so must subtract the total owner

xfed <- mtawage3 |> 
  filter(ind=="allind" & owncode=="fedgov")
xfed |> filter(region=="nys", date==max(date))
xfed |> filter(region=="nyc", date==max(date))

count(xk12, area)


mtabase1 <- mtawage3 |> 
  filter(ind=="allind", owncode=="total") |> 
  select(-c(ind, owncode)) |> 
  rename(total=value) |> 
  left_join(xk12all |> 
              filter(owncode=="total") |> # use this due to data limitations
              select(date, area, name, xk12=value),
            by = join_by(date, area, name)) |> 
  left_join(xfed |> 
              select(date, area, name, xfed=value),
            by = join_by(date, area, name)) |> 
  mutate(pmtbase=total - naz(xk12) - naz(xfed))

count(mtabase1, area, region)
skim(mtabase1)
glimpse(mtabase1)

saveRDS(mtabase1, here::here("data", "mta_wagebase.rds"))


```

Here is a summary of the resulting data file.

```{r}
#| label: summary
#| eval: false
#| include: true

mtabase1 <- readRDS(here::here("data", "mta_wagebase.rds"))
skim(mtabase1)

```

And correlation between total wages and the tax base after removing federal wages and K12 wages:

```{r}
#| label: correlation
#| eval: false
#| include: true

mtabase1 |> filter(name=="totwage") |> select(total, pmtbase) |> cor()

```

```{r}
#| label: explore
#| eval: false
#| include: false

mtabase1 <- readRDS(here::here("data", "mta_wagebase.rds"))

count(mtabase1, region)

mtabase1 |> filter(name=="totwage") |> select(total, pmtbase) |> cor()

mtabase1 |> 
  filter(name=="totwage") |> 
  summarise(total=sum(total), .by=c(date, region)) |> 
  ggplot(aes(date, total)) +
  geom_line() +
  geom_point() +
  geom_smooth() +
  scale_x_date(breaks = "6 months", date_labels = "%Y-%b") +
  facet_wrap(~region, scales="free_y", ncol=1) +
  theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust=0.5)) 


mtabase1 |> 
  filter(name=="totwage") |> 
  summarise(total=sum(total), .by=c(date, region)) |> 
  arrange(date) |> 
  mutate(pch=total / lag(total) - 1, .by=region) |> 
  mutate(region=factor(region, levels=c("nyc", "suburban", "nys"))) |> 
  filter(year(date) >= 2010) |> 
  ggplot(aes(date, pch)) +
  geom_line() +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0) +
  scale_x_date(name=NULL, breaks = "3 months",  date_labels = "%Y-%b") +
  scale_y_continuous(name="% change from year ago", 
                     breaks=seq(-.5, .5, .05),
                     labels=label_percent(accuracy=1)) +
  facet_wrap(~region, ncol=1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust=0.5))


```

```{r}
#| label: levels
#| eval: false
#| include: true
#| fig-width: 8
#| fig-height: 10

mtabase1 <- readRDS(here::here("data", "mta_wagebase.rds"))

# count(mtabase1, region)

pdata1 <- mtabase1 |> 
  filter(name=="totwage") |> 
  summarise(pmtbase=sum(pmtbase), .by=c(date, region)) |> 
  arrange(date)

mtasum <- pdata1 |> 
  filter(region %in% c("nyc", "suburban")) |> 
  summarise(pmtbase=sum(pmtbase), .by=c(date)) |> 
  mutate(region="mta")

pdata <- bind_rows(pdata1, mtasum) |>
  mutate(region=factor(region, levels=c("nyc", "suburban", "mta", "nys")))


brks <- seq.Date(from=as.Date("2000-01-01"),
                 to=as.Date("2030-01-01"),
                 by="3 months")

pdata |> 
  filter(year(date) >= 2015, region!="nys") |> 
  ggplot(aes(date, pmtbase)) +
  geom_line() +
  geom_point() +
  geom_smooth() +
  scale_x_date(name=NULL, breaks = brks,  date_labels = "%Y-%b") +
  scale_y_continuous(name="amount ($billions)", 
                     labels=scales::label_comma(accuracy=1, scale=1e-9)) +
  facet_wrap(~region, ncol=1, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust=0.5))

```

```{r}
#| label: growth-rates
#| eval: false
#| include: true
#| fig-width: 8
#| fig-height: 10

mtabase1 <- readRDS(here::here("data", "mta_wagebase.rds"))

# count(mtabase1, region)

pdata1 <- mtabase1 |> 
  filter(name=="totwage") |> 
  summarise(pmtbase=sum(pmtbase), .by=c(date, region)) |> 
  arrange(date)

mtasum <- pdata1 |> 
  filter(region %in% c("nyc", "suburban")) |> 
  summarise(pmtbase=sum(pmtbase), .by=c(date)) |> 
  mutate(region="mta")

pdata <- bind_rows(pdata1, mtasum) |>  
  mutate(pch=pmtbase / lag(pmtbase) - 1, .by=region) |> 
  mutate(region=factor(region, levels=c("nyc", "suburban", "mta", "nys")))

brks <- seq.Date(from=as.Date("2000-01-01"),
                 to=as.Date("2030-01-01"),
                 by="3 months")

pdata |> 
  filter(year(date) >= 2015, region!="nys") |> 
  ggplot(aes(date, pch)) +
  geom_line() +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0) +
  scale_x_date(name=NULL, breaks = brks,  date_labels = "%Y-%b") +
  scale_y_continuous(name="% change from year ago", 
                     breaks=seq(-.5, .5, .05),
                     labels=label_percent(accuracy=1)) +
  facet_wrap(~region, ncol=1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust=0.5))


```


```{r}
#| label: unknown
#| eval: false

# Example date
date_example <- as.Date("2023-05-15")

# Find the first day of the quarter
first_day_of_quarter <- floor_date(date_example, "quarter")
first_day_of_quarter


pmtfixed <- readRDS(here::here("data", "mta_pmt_collections.rds"))
count(pmtfixed, vname)

wtax <- pmtfixed |> 
  filter(vname=="pmt_wage", date >= "2009-10-01") |> 
  select(date, wagetax=value) |> 
  mutate(date=floor_date(date, "quarter")) |> 
  summarise(wagetax=sum(wagetax), .by=date) |> 
  arrange(date) |> 
  mutate(fgroup="wagetax", pchya=wagetax / lag(wagetax, 4) - 1)

comp <- bind_rows(ihs3,
                  wtax |> select(date, value=wagetax, fgroup, pchya))
    
p <- comp |> 
  filter(year(date)>=2015, date <= "2023-04-01") |> 
  ggplot(aes(date, pchya, colour=fgroup)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_y_continuous(breaks=seq(-.3, .3, .01), labels = scales::percent_format(accuracy = .1)) +
  ggtitle("MTA region year-over-year growth in seasonally adjusted and non-adjusted wages") +
  theme_bw()
p  
ggsave(path(respath, "ns_vs_sa_wages.png"), plot=p)  


```




## Documentation

### QCEW data

```{=html}
<!-- links

Starting page:
  https://dol.ny.gov/occupational-and-industry-data

To get data:
  go to the bottom of this page
    https://dol.ny.gov/quarterly-census-employment-and-wages
  where it says "More Information" and includes the statement:
  
  "The entire set of currently available NAICS-based QCEW data is available by downloading the file qcew.zip. Note: Prior to 2018, the data in this file is available down to the three-digit NAICS level. For 2018 going forward, the data in this file is available down to the six-digit level."

  Then download the qcew.zip file.
  
  https://statistics.labor.ny.gov/qcew.zip  


From the README: 
DATA ELEMENTS AND DEFINITIONS for qcew_quarterly.txt
Data Element    Definition
AREATYPE        Type of Area (State, MSA, Labor Market Region, LWIA, County)
AREA            Area Name
YEAR            Year
QUARTER         Quarter of the year (1,2,3,4)
NAICS           NAICS industry classification code
NAICS_LEVEL           NAICS industry level
TITLE           NAICS Title
ESTAB           Number of establishments
MNTH1EMP        Number employed for first month of given quarter
MNTH2EMP        Number employed in second month
MNTH3EMP        Number employed in third month
TOTWAGE     Total wages
OWNER   Private or Government

New York City employment
https://www.nyc.gov/assets/omb/downloads/csv/fy2025/nycemploy-nsa08-24.csv


-->
```


```{r}
#| label: docs-of-interest
#| include: false

# See NYC Comptroller Brad Lander's monthly report on the NYC economy:
#   https://comptroller.nyc.gov/newsroom/newsletter/


```


```{r}
#| label: vroom-zip-notes

# exdir <- tempfile()
# system.time(zip::unzip(zpath, files=csvfiles, exdir = exdir)) # 1.8 secs
# system.time(df <- vroom(purrr::map(csvfiles, \(x)  unz(zpath, x)))) # 2.8 secs
# df <- vroom(purrr::map(csvfiles, \(x)  unz(zpath, x)))
# glimpse(df)
# system.time(zip::unzip(zpath, exdir = exdir)) # 9.1 secs
# system.time(utils::unzip(zpath, exdir = exdir)) # 8.2 secs
# utils::unzip(zpath, files = "qcew_annual_2000.csv", exdir = exdir)

# con1 <- unz(zpath, csv_files[1])
# con1
# open(con1)
# df <- vroom(con1)
# close(con1)

# Use vroom to read all CSV files in the temporary directory
# library(vroom)
# file_paths <- file.path(temp_dir, csv_files)  # Paths to the extracted CSV files
# data_list <- lapply(file_paths, vroom)

```


## NYC employment data

```{r}
#| label: ONETIME-get-save-nyc-employment
#| eval: false

nycurl <- "https://www.nyc.gov/assets/omb/downloads/csv/fy2025/nycemploy-nsa08-24.csv"
fname <- fs::path_file(nycurl)
fpath <- here::here("data", "nycomb", fname)
download.file(nycurl, destfile = fpath, mode="wb")

df <- read_csv(fpath, skip=2)
glimpse(df)

df2 <- df |> 
  filter(row_number() != 1) |> 
  select(date=1, total=2, private=3, govt=29) |> 
  mutate(date = ymd(paste0(substr(date, 1, 4), "-", substr(date, 6, 7), "-01")),
         across(-date, as.numeric)) 

df2 |> 
  pivot_longer(-date) |> 
  mutate(pchya=value / lag(value, 12) - 1) |>
  filter(year(date) >= 2021) |> 
  ggplot(aes(date, pchya, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  # geom_smooth() +
  scale_x_date(name=NULL, breaks = "6 months") +
  scale_y_continuous(name="% change from year ago", 
                     breaks=seq(-.3, .3, .02),
                     labels = scales::percent_format(accuracy = .1)) +
  # facet_wrap(~"NYC employment", scales="free_y", ncol=1) +
  theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust=0.5))

df2 |> 
  pivot_longer(-date) |> 
  mutate(pchya=value / lag(value, 12) - 1) |>
  filter(year(date) >= 2021) |> 
  filter(name=="total") |>
  ggplot(aes(date, pchya)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  geom_smooth() +
  scale_x_date(name=NULL, breaks = "3 months") +
  scale_y_continuous(name="% change from year ago", 
                     breaks=seq(-.3, .3, .02),
                     labels = scales::percent_format(accuracy = .1)) +
  theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust=0.5))


# seasonally adjusted
saurl <- "https://www.nyc.gov/assets/omb/downloads/csv/fy2025/nycemploy-sa08-24.csv"
sadf <- read_csv(saurl, skip=3) |> 
  filter(row_number() != 1) |> 
  select(date=1, total=2, private=3, govt=29) |> 
  mutate(date = ymd(paste0(substr(date, 1, 4), "-", substr(date, 6, 7), "-01")),
         across(-date, as.numeric)) 

sadf |> 
  select(date, total) |>
  mutate(pchya=total / lag(total, 12) - 1) |>
  # filter(year(date) >= 2023) |> 
  filter(date >= "2023-07-01") |>
  ggplot(aes(date, pchya)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  geom_smooth() +
  scale_x_date(name=NULL, breaks = "1 month") +
  scale_y_continuous(name="% change from year ago", 
                     breaks=seq(-.3, .3, .005),
                     labels = scales::percent_format(accuracy = .1)) +
  ggtitle("NYC seasonally adjusted employment year-over-year % growth",
          subtitle = "Source: NYC OMB (https://www.nyc.gov/assets/omb/downloads/csv/fy2025/nycemploy-sa08-24.csv)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust=0.5)) 

sadf |> 
  select(date, total) |>
  filter(year(date) >= 2023) |> 
  ggplot(aes(date, total)) +
  geom_line() +
  geom_point() +
  geom_smooth() +
  scale_x_date(name=NULL, breaks = "1 month") +
  scale_y_continuous(name="% change from year ago", 
                     breaks=seq(-.3, .3, .005),
                     labels = scales::percent_format(accuracy = .1)) +
  ggtitle("NYC seasonally adjusted employment year-over-year % growth",
          subtitle = "Source: NYC OMB (https://www.nyc.gov/assets/omb/downloads/csv/fy2025/nycemploy-sa08-24.csv)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust=0.5)) 
  
```
