# PMT alternative policies

```{r}
#| label: setup

suppressPackageStartupMessages(source(here::here("R", "libraries.r")))
suppressPackageStartupMessages(source(here::here("R", "libraries_ts.r")))

source(here::here("R", "constants.r"))
source(here::here("R", "functions.r"))
source(here::here("pmt", "pmt_constants_and_functions.r"))

# GLOBAL OPTIONS:
#  eval: true
#  output: false
#  echo: false

```

```{r}
#| label: prepare-pmt-data
#| output: false

pmtbl <- readRDS(fs::path(PDINTERMEDIATE, "pmt2024_baseline.rds"))
glimpse(pmtbl)
count(pmtbl, status)

# define payroll cuts and labels
(payroll_cuts <- c(0, 500e3, seq(1e6, 2e6, 200e3), seq(2e6, 5e6, 1e6), Inf, 1.25e6, 1.75e6) |> unique() |> sort())
paygroup_labels <- cutlabs(payroll_cuts)

pmtbl2 <- pmtbl |> 
  mutate(paygroup=cut(esize_payestab, payroll_cuts, right=FALSE, labels=FALSE),
         paylabel=getlabel(paygroup, paygroup_labels)) |> 
  arrange(paygroup)


```

## Raise taxability threshold from \$1.2m to \$1.6m and raise top rate proportionately in NYC & suburbs to maintain revenue-neutrality, 2024

```{r}
#| label: calc-revenue-neutral-rate
#| output: false

base_rates

raise_exclusion <- base_rates |> 
  mutate(payroll_ub = ifelse(payroll_ub==1.25e6, 1.6e6, payroll_ub)) |> # raise the bottom
  filter(payroll_ub != 1.5e6)

xup <- pmtbl2 |> 
  mutate(get_info(esize_payestab, nyc, rates=raise_exclusion, prefix="xup_"),
         xup_tax= esize_payroll * xup_rate)

xup |> 
  mutate(paygroup = ifelse(esize_payestab >= 1.75e6, 99, paygroup),
         paylabel=ifelse(paygroup==99, "top", paylabel)) |> 
  summarise(n=n(),
          across(c(base_tax, xup_tax),
                 \(x) sum(x)),
                 .by=c(paygroup, paylabel)) |> 
  janitor::adorn_totals() |> 
  mutate(change=xup_tax - base_tax) |> 
  knitr::kable(decimals=1, format.args = list(big.mark = ","))
# 1.7 % rate increase needed


adjust <- 1.016996
raise_exclusion_and_top_rate <- base_rates |> 
  mutate(payroll_ub = ifelse(payroll_ub==1.25e6, 1.6e6, payroll_ub)) |> # raise the bottom
  mutate(nyc = ifelse(is.infinite(payroll_ub), nyc * adjust, nyc),
         suburbs = ifelse(is.infinite(payroll_ub), suburbs * adjust, suburbs)) |> 
  mutate(across(c(nyc, suburbs), \(x) round(x, 5))) |> 
  filter(payroll_ub != 1.5e6)

raise_exclusion_and_top_rate |> knitr::kable()

xuptopup <- pmtbl2 |> 
  mutate(get_info(esize_payestab, nyc, rates=raise_exclusion_and_top_rate, prefix="xuptopup_"),
         xuptopup_tax= esize_payroll * xuptopup_rate)

xuptopup |>
  summarise(n=n(),
          across(c(base_tax, xuptopup_tax),
                 \(x) sum(x)),
                 .by=c(paygroup, paylabel)) |> 
  janitor::adorn_totals() |> 
  mutate(change=xuptopup_tax - base_tax) |> 
  knitr::kable(decimals=1, format.args = list(big.mark = ","))

```

```{r}
#| label: get-rev-neutral-rates

data1 <- xuptopup |> 
  summarise(n=n(),
            across(c(esize_estabsadj, 
                     esize_emp,
                     # esize_payroll,
                     base_tax,
                     xuptopup_tax),
                   \(x) sum(x)),
                   .by=c(nyc, paygroup, paylabel))

# data1

tabdata <- data1 |> 
  select(-n, -esize_emp) |> 
  mutate(nyc=factor(nyc, levels=c(TRUE, FALSE), labels = c("NYC", "Suburbs"))) |> 
  pivot_wider(names_from=nyc, values_from = c(esize_estabsadj, base_tax, xuptopup_tax)) |> 
  mutate(base_total_tax = base_tax_NYC + base_tax_Suburbs,
         alt_total_tax = xuptopup_tax_NYC + xuptopup_tax_Suburbs) |> 
  arrange(paygroup) |> 
  select(paylabel, contains("NYC"), contains("Suburbs"), base_total_tax, alt_total_tax) |> 
  janitor::adorn_totals() |> 
  mutate(change=alt_total_tax - base_total_tax,
         nyc_change_share = (xuptopup_tax_NYC - base_tax_NYC) / change,
        nyc_change_share = ifelse(paylabel=="Total", NA_real_, nyc_change_share))
# tabdata

```

```{r}
#| label: show-tax-rates
#| output: true

base_rates |> 
  gt() |> 
  tab_header("Current PMT rates") |> 
  cols_label(payroll_ub = "Payroll upper bound") |> 
  fmt_number(columns = payroll_ub, decimal=0) |> 
  fmt_percent(columns = c(nyc, suburbs), decimals=3)

 
raise_exclusion_and_top_rate |> 
  gt() |> 
  tab_header("Revenue neutral rates, increasing exclusion and top rate") |> 
  cols_label(payroll_ub = "Payroll upper bound") |> 
  fmt_number(columns = payroll_ub, decimal=0) |> 
  fmt_percent(columns = c(nyc, suburbs), decimals=3)

```

```{r}
#| label: examine-revenue-neutral
#| output: true
#| column: page

tabdata |> 
  gt() |> 
  tab_header("Revenue-neutral increase in tax-exclusion threshold (to $1.6m annual from $1.25m) with proportionate increase in top rate (+1.7%)",
             subtitle = "Private sector only. Amounts in $ millions") |> 
  cols_label(paylabel="Average establishment payroll ($)",
             contains("estabsadj") ~ "# of establishments",
             contains("base_tax") ~ "Baseline PMT wage tax ($m)",
             contains("xuptopup_tax") ~ "Alternative PMT wage tax ($m)",
             base_total_tax = "Baseline total PMT wage tax ($m)",
             alt_total_tax = "Alternative total tax ($m)",
             change = "Change in tax ($m)",
             nyc_change_share = "NYC share of tax change") |> 
  tab_spanner(label="New York City", columns=contains("NYC", ignore.case = FALSE)) |> 
  tab_spanner(label="Suburbs", columns=contains("suburbs", ignore.case = TRUE)) |> 
  fmt_number(columns = contains("estabsadj"), decimals=0) |>
  fmt_number(columns = c(contains("tax"), change), scale = 1e-6, decimals = 1) |>
  fmt_currency(rows=c(1, nrow(tabdata)), columns = c(contains("tax"), change), scale=1e-6, decimals=1) |> 
  fmt_percent(columns = c(nyc_change_share), decimals=1)  |> 
  sub_missing(missing_text = "--") |> 
  tab_source_note(source_note = source_note) |> 
  tab_source_note(source_note = long_note)

```

## Increase revenue by \$500 million while raising taxability threshold to \$1.6 million

```{r}
#| label: determine-500m-rates

adjust <- 1.216

raise_halfbillion_and_increase_exclusion <- base_rates |> 
  mutate(payroll_ub = ifelse(payroll_ub==1.25e6, 1.6e6, payroll_ub)) |> # raise the bottom threshold
  mutate(nyc = ifelse(is.infinite(payroll_ub), nyc * adjust, nyc),
         suburbs = ifelse(is.infinite(payroll_ub), suburbs * adjust, suburbs)) |> 
  mutate(across(c(nyc, suburbs), \(x) round(x, 5))) |> 
  filter(payroll_ub != 1.5e6)

raise_halfbillion_and_increase_exclusion |> knitr::kable()

alt <- pmtbl2 |> 
  mutate(get_info(esize_payestab, nyc, rates=raise_halfbillion_and_increase_exclusion, prefix="alt_"),
         alt_tax= esize_payroll * alt_rate)

alt |>
  summarise(n=n(),
          across(c(base_tax, alt_tax),
                 \(x) sum(x)),
                 .by=c(paygroup, paylabel)) |> 
  janitor::adorn_totals() |> 
  mutate(change=alt_tax - base_tax) |> 
  knitr::kable(decimals=1, format.args = list(big.mark = ","))

```

```{r}
#| label: show-increase-tax-rates
#| output: true

base_rates |> 
  gt() |> 
  tab_header("Current PMT rates") |> 
  cols_label(payroll_ub = "Payroll upper bound") |> 
  fmt_number(columns = payroll_ub, decimal=0) |> 
  fmt_percent(columns = c(nyc, suburbs), decimals=3)

 
raise_halfbillion_and_increase_exclusion |> 
  gt() |> 
  tab_header("Rates to raise $500 million: Combination of higher top rate and increase in exclusion from tax") |> 
  cols_label(payroll_ub = "Payroll upper bound") |> 
  fmt_number(columns = payroll_ub, decimal=0) |> 
  fmt_percent(columns = c(nyc, suburbs), decimals=3)

```

```{r}
#| label: calc-increase

data1 <- alt |> 
  summarise(n=n(),
            across(c(esize_estabsadj, 
                     esize_emp,
                     base_tax,
                     alt_tax),
                   \(x) sum(x)),
                   .by=c(nyc, paygroup, paylabel))

tabdata <- data1 |> 
  select(-n, -esize_emp) |> 
  mutate(nyc=factor(nyc, levels=c(TRUE, FALSE), labels = c("NYC", "Suburbs"))) |> 
  pivot_wider(names_from=nyc, values_from = c(esize_estabsadj, base_tax, alt_tax)) |> 
  mutate(base_total_tax = base_tax_NYC + base_tax_Suburbs,
         alt_total_tax = alt_tax_NYC + alt_tax_Suburbs) |> 
  arrange(paygroup) |> 
  select(paylabel, contains("NYC"), contains("Suburbs"), base_total_tax, alt_total_tax) |> 
  janitor::adorn_totals() |> 
  mutate(change=alt_total_tax - base_total_tax,
         nyc_change_share = (alt_tax_NYC - base_tax_NYC) / change)
  #      nyc_change_share = ifelse(paylabel=="Total", NA_real_, nyc_change_share))
# tabdata

```

```{r}
#| label: show-500-million-increase
#| output: true
#| column: page

tabdata |> 
  gt() |> 
  tab_header("$500 million increase from 21.6% increase in top rate and raising tax-exclusion threshold to $1.6m annual from $1.25m",
             subtitle = "Private sector only. Amounts in $ millions") |> 
  cols_label(paylabel="Average establishment payroll ($)",
             contains("estabsadj") ~ "# of establishments",
             contains("base_tax") ~ "Baseline PMT wage tax ($m)",
             contains("alt_tax") ~ "Alternative PMT wage tax ($m)",
             base_total_tax = "Baseline total PMT wage tax ($m)",
             alt_total_tax = "Alternative total tax ($m)",
             change = "Change in tax ($m)",
             nyc_change_share = "NYC share of tax change") |> 
  tab_spanner(label="New York City", columns=contains("NYC", ignore.case = FALSE)) |> 
  tab_spanner(label="Suburbs", columns=contains("suburbs", ignore.case = TRUE)) |> 
  fmt_number(columns = contains("estabsadj"), decimals=0) |>
  fmt_number(columns = c(contains("tax"), change), scale = 1e-6, decimals = 1) |>
  fmt_currency(rows=c(1, nrow(tabdata)), columns = c(contains("tax"), change), scale=1e-6, decimals=1) |> 
  fmt_percent(columns = c(nyc_change_share), decimals=1)  |> 
  sub_missing(missing_text = "--") |> 
  tab_source_note(source_note = source_note) |> 
  tab_source_note(source_note = long_note)

```

## Alternative MTA region tax (\$500 million increase and taxability threshold raised to \_\$1.6 million) by firm size and county -- Excel file

This section writes county-by-county results to an Excel file.

The columns are:

-   `area`
-   `paygroup`
-   `paylabel`
-   `esize_estabsadj` -- number of establishments in the county-payroll group
-   `esize_emp` -- number of employees in the group
-   `base_tax` -- base (current law) tax in the group at 2024 levels
-   `alt_tax` -- alternative-policy tax in the group at 2024 levels

Note that the file does not have totals by county or by payroll size.

```{r}
#| label: tax-mta-county
#| eval: true
#| output: true
#| column: page

data1 <- alt |> 
  mutate(area = ifelse(nyc, "New York City", area)) |> 
  summarise(across(c(esize_estabsadj, 
                     esize_emp,
                     base_tax,
                     alt_tax),
                   \(x) sum(x)),
                   .by=c(area, paygroup, paylabel))

openxlsx::write.xlsx(data1, file=fs::path(PDMTA, "increase500m_by_county_and_establishment_payroll.xlsx"))

```

[Download Excel File -- \$500m increase in tax at 2024 levels, by county and payroll size](data/for_mta/increase500m_by_county_and_establishment_payroll.xlsx)
