---
output: html_document
editor_options: 
  chunk_output_type: console
---

# PMT model

```{r}
#| label: setup

suppressPackageStartupMessages(source(here::here("R", "libraries.r")))
suppressPackageStartupMessages(source(here::here("R", "libraries_ts.r")))

source(here::here("R", "constants.r"))
source(here::here("R", "functions.r"))
source(here::here("pmt", "pmt_constants_and_functions.r"))

# GLOBAL OPTIONS:
#  eval: true
#  output: false
#  echo: false

```

## Prepare PMT database for 2024

This section prepares and saves a 2024 database with current-law tax. It only needs to be rerun if underlying data change or 2024 assumptions change.

It does the following:

-   Get 2023 base data previously constructed from QCEW and CBP data
-   Update data to 2024 based on wage and employment growth estimates
-   Calculate 2024 tax, examine and possibly calibrate
-   Save the 2024 database

Later sections use the saved database.

```{r}
#| label: get-data
#| eval: false

pmtdb1 <- readRDS(fs::path(PDINTERMEDIATE, "pmt_final.rds"))
glimpse(pmtdb1)
head(pmtdb1)

uqmta <- readRDS(fs::path(PDINTERMEDIATE, "uqmta.rds")) # non-totals qcew
qcew_tots <- uqmta |> 
  summarise(n = n(),
            across(c(estabs, emp, payroll), sum),
            .by=c(owner, status)) |> 
  mutate(avgpay = payroll / emp,
         avgpayroll = payroll / estabs)
qcew_tots

```

```{r}
#| label: forecast-underlying-data
#| eval: false

# first forecast to 2024 and calibrate
# growth rates emp 1.9 wage rate 2.1 nyc council https://council.nyc.gov/wp-content/uploads/2024/03/Dashboard_FY25-Prelim_EMBARGOED.pdf
# that was before start of the year chatbots suggest higher wage growth

# 2023 to 2024 growth rates
empgrowth <- .02
wagerategrowth <- .04

# first, update base data to 2024 and ensure internal consistency
# grow all employment items including range endpoints
# do NOT grow number of establishments
# grow total payroll
pmtdb2 <- pmtdb1 |> 
  select(year, owner, area, nyc, naics, qtitle, status, estabs, emp, payroll, esize, elb, eub, esize_estabsadj, esize_avgemp) |> 
  mutate(year=2024,
         across(c(emp, elb, eub, esize_avgemp),
                \(x) x * (1 + empgrowth)),
         payroll = payroll * (1 + empgrowth) * (1 + wagerategrowth))

# check totals variables; also compare 2023 to raw qcew data
bind_rows(pmtdb1 |> mutate(src="y2023"),
          pmtdb2 |> mutate(src="y2024")) |> 
  summarise(payroll=first(payroll),
            emp=first(emp),
            .by=c(src, area, naics)) |> 
  summarise(payroll=sum(payroll),
            emp=sum(emp), 
            .by=src) |> 
  mutate(avgpay=payroll / emp) |> 
  pivot_longer(-src) |> 
  pivot_wider(names_from = src) |> 
  mutate(pch=y2024 / y2023 - 1)

qcew_tots

# check details variables
bind_rows(pmtdb1 |> mutate(src="y2023"),
          pmtdb2 |> mutate(src="y2024")) |>
  mutate(emp = esize_estabsadj * esize_avgemp) |> 
  summarise(emp=sum(emp), .by=src) |> 
  pivot_longer(cols = emp) |> 
  pivot_wider(names_from = src) |> 
  mutate(pch=y2024 / y2023 - 1)

qcew_tots

pmtdb3 <- pmtdb2 |> 
  mutate(esize_emp = esize_estabsadj * esize_avgemp, # number of estabs * avgemp
         esize_avgpay = payroll / emp, # same for everyone
         esize_payroll = esize_avgpay * esize_emp,
         esize_payestab = esize_payroll / esize_estabsadj)

```

```{r}
#| label: calc-2024-tax-and-calibrate-and-save
#| eval: false

# wage tax per tax dept monthly collections, 2024 -- $2,960.405 million
# target_tax_2024 <- 2960.405 * 1e6 # my Nov 2024 forecast
target_tax_2024 <- 2924.692 * 1e6 # actual Oct 2023 - Sep 2024
# target_tax_2024 <- 2924.692 * 1e6 * .95 # if 5% of total is late payments

taxdata_raw <- pmtdb3 |> 
  # drop any records we created where there are zero establishments as CBP would not have shown that
  # however, we don't drop records where there are establishments but zero employment because CBP data might have shown that
  filter(esize_estabsadj != 0) |> # PERHAPS DO THIS EARLIER, IN FILE CREATION, RATHER THAN HERE??? or maybe this is best
  # esize_payestab is average pay-per-establishment in a given establishment size group (defined by # of employees per establishment)
  # esize_payroll is total payroll in the group
  mutate(get_info(esize_payestab, nyc, rates=base_rates),
         base_tax= esize_payroll * base_rate)

# calculate total tax and calibrate to target
taxraw_tots <- taxdata_raw |> 
  mutate(nestabs_taxable = esize_estabsadj * (base_tax > 0),
         emp_taxable = esize_emp * (base_tax > 0),
         payroll_taxable = esize_payroll  * (base_tax > 0)) |> 
  summarise(across(c(nestabs_taxable, emp_taxable, payroll_taxable, esize_estabsadj, esize_emp, esize_payroll, base_tax), sum)) |> 
  mutate(avgpay=payroll_taxable / emp_taxable,
         base_etr = base_tax / payroll_taxable)
  
taxraw_tots # $2.567 billion calculated vs. ~$2.925 billion from collections -- about $358m
# perhaps $200m could come from that?
qcew_tots

# determine adjustment -- about 8-15% upward adjustment depending on 2024 assumption -- we'll do entirely through payroll?
# what would cause this? 
#   all 12-month periods near 2024 have ~$2.9 to $2.95 billion; we are short maybe $350 million
#   could collections reflect late payments?
#   are some government entities taxable? yes -- maybe $100-200 of that could be govts, maybe rest could be distrib of firm size
#   could firms be larger than establishments so that tax rates are higher than I use?
#   could I need to distribute establishments better giving more larger establishments
# (payroll_adjust <- target_tax_2024 / taxraw_tots$tax) # 13.9% with my favored adjustment

# according to chatgpt: 
  # State and Local Government Entities: The MCTMT does not explicitly exempt
  # state or local government entities from its provisions. Therefore, unless
  # another specific exemption applies, these entities are subject to the tax if
  # they meet the general criteria.

# SOLUTION: DO NOT adjust details, but adjust at the end if needed, later maybe revisit distribution of establishments
# 
# save baseline tax for 2024 ----

# mutate(get_info(esize_payestab, nyc, rates=alt_rates, prefix="alt_")) |> 

saveRDS(taxdata_raw, fs::path(PDINTERMEDIATE, "pmt2024_baseline.rds"))

```

## PMT model data notes

The PMT database used below has estimates of \# of establishments, \# of employees, and total payroll by county, industry, and establishment size (# of employees). It was constructed by (1) starting with QCEW data on establishments, employees, and wages by county and industry, and (2) using establishment-size information from County Business Patterns to estimate the breakdown of the QCEW by establishment size.

An establishment is a workplace.

The database has several weaknesses:

-   The data are based on establishment size but PMT tax rates are based on business size (the size of the firm in the MTA region). Thus, tax rates will tend to be understated somewhat because firms are bigger than establishments in the case of multi-establishment firms.
-   The data do not include government employers. In general, New York state government employers and New York local non-school-district government employers are taxable. This could understate tax revenue by 5% (+/-).
-   The data do not include self-employed individuals who are subject to the net self-employment earnings tax that is a counterpart to the payroll tax. External adjustments should be made to accounty for this.

I can improve the database to adjust for the distinction between firm size and establishment size, and to add in data for government employers, but I have not done so yet.

Benchmarks from tax collections for the 12 months ending Sept 2024, based on Tax Department OpenData:

-   Payroll tax collections were \$2,924 million
-   Net earnings self-employment tax collections were \$140 million

```{r stop_here, echo=FALSE}
knitr::knit_exit()
```
