---
output: html_document
editor_options: 
  chunk_output_type: console
---

{{< include _setup.qmd >}}

# Add \# of firms relative to \# of establishments

Use data from SUSB to estimate the number of firms in each county-industry-(establishment-size) cell in the PMT database.

Steps:

-   From SUSB, create a lookup table with average \# of firms per establishment by 2-digit industry and industry-specific establishment size group (but not by county -- the presumption, based on examination of the SUSB data, is that \# of firms per establishment varies greatly by establishment size within industry, but less so by county)
-   For our QCEW-CBP-based PMT database, calculate average establishment size for each cell (county, industry, establishmnt-size)
-   Match the average establishment size of each record in the PMT database against the SUSB lookup table to get the industry-average number of firms per establishment for the corresponding establishment-size group
    -   For PMT cells that do not have an industry-specific match in the SUSB lookup table, use the all-industry (snaics == "--") values
-   In the PMT database, multiply industry-average establishment-size-group \# of firms per establishment pulled from the SUSB lookup table by \# of establishments to get estimated \# of firms for each cell in the PMT database
-   In the PMT database, we can now divide total payroll for a cell by estimated number of firms in the cell to get average payroll per firm. This can be used to get the tax rate for the payroll in that cell.

```{r}
#| label: get-data
#| output: false

pmtdb1 <- readRDS(fs::path(PDINTERMEDIATE, "qcew_allocated.rds"))
susb <- readRDS(fs::path(PDINTERMEDIATE, "susb2021_mta.rds"))

```

## Prepare SUSB lookup table

```{r}
#| label: create-susb-lookup-table
#| output: false

glimpse(susb)

# get firms per establishment -- s prefix indicates the data are from susb
firms_per_estabs <- susb |> 
  select(area, snaics, stitle, firmsize, sfirms, sestabs, semp) |> 
  mutate(sfirms_estabs = sfirms / sestabs,
         semp_estabs = semp / sestabs, 
         semp_firms = semp / sfirms,
         sestabs_firms = sestabs / sfirms)

# get industries for match: 2-digit and dual-2-digit (e.g., 31-33, 44-45, 48-49) ----
firms_per_estabs_match1 <- firms_per_estabs |> 
  filter(nchar(snaics) == 2 |
           str_detect(snaics, "-"))

firms_per_estabs_match1 |> 
  distinct(snaics, stitle) |> 
  arrange(snaics) # we have 21 2-digit-like industry groups

# take a quick look
summary(firms_per_estabs_match1) # nothing missing

firms_per_estabs_match1 |> 
  summarise(n=n(),
            sfirms_estabs=median(sfirms_estabs),
            sestabs_firms=median(sestabs_firms), # the inverse
            .by=firmsize) # at higher firm sizes we have more estabs per firm


# create the look up table for firms per establishment
susb_lookup_fpe <- firms_per_estabs_match1 |> 
  mutate(esize_group = ntile(semp_estabs, 4), # divide the records into 4 establishment-size groups for each industry
         .by=c(snaics, stitle)) |> 
  select(snaics, stitle, esize_group, semp_estabs, sfirms_estabs) |> 
  summarise(
    semp_estabs_min = min(semp_estabs),
    semp_estabs_max = max(semp_estabs),
    sfirms_estabs_min = min(sfirms_estabs),
    sfirms_estabs_mdn = median(sfirms_estabs),
    sfirms_estabs_max = max(sfirms_estabs),
    .by=c(snaics, stitle, esize_group)) |> 
  arrange(snaics, esize_group) |> 
  # get industry-specific cutpoints for establishment-size groups (# of employees per establishment)
  mutate(lowcut = case_when(esize_group == 1 ~ 0,
                          esize_group > 1 ~ pmean(semp_estabs_min, lag(semp_estabs_max)),
                          .default = 0),
         hicut = case_when(esize_group == 4 ~ Inf,
                           esize_group < 4 ~ pmean(semp_estabs_max, lead(semp_estabs_min)),
                           .default = Inf)) |> 
  rename(naics_match = snaics) |> 
  arrange(naics_match, esize_group)

susb_lookup_fpe

susb_lookup_fpe_inds <- count(susb_lookup_fpe, naics_match, stitle) # for checking against

```

## Prepare PMT database to match against SUSB lookup table

Create a variable that will match against naics_match in the lookup table.

```{r}
#| label: prepare-pmtdb-with-matching-variable
#| output: false

pmtdb_prep1 <- pmtdb1 |> 
  mutate(naics2 = str_sub(naics, 1, 2),
         naics_match = case_when(
           naics2 %in% c("31", "32", "33") ~ "31-33",
           naics2 %in% c("44", "45") ~ "44-45",
           naics2 %in% c("48", "49") ~ "48-49",
           .default = naics2)) 

check <- count(pmtdb_prep1, naics_match, naics2, naics, title) |> 
  left_join(susb_lookup_fpe_inds |>
              select(naics_match, stitle),
            by = join_by(naics_match))

pmtdb_prep2 <- pmtdb_prep1 |> 
  mutate(naics_match = ifelse(!naics_match %in% susb_lookup_fpe_inds$naics_match,
                              "--",
                              naics_match))

check <- count(pmtdb_prep2, naics_match, naics2, naics, title) |> 
  left_join(susb_lookup_fpe_inds |>
              select(naics_match, stitle),
            by = join_by(naics_match)) # we now have matches for everything

```

## Match pmtdb and SUSB lookup to get firm-establishment ratios for corresponding industry and establishment size

```{r}
#| label: match-pmtdb-to-susb-lookup
#| output: false

summary(pmtdb_prep2)

pmtdb_match <- pmtdb_prep2 |> 
  left_join(susb_lookup_fpe |> 
              select(naics_match, lowcut, hicut, fpe = sfirms_estabs_mdn),
            join_by(naics_match, gemp_est >= lowcut, gemp_est <= hicut))
summary(pmtdb_match)  

pmtdb_match2 <- pmtdb_match |> 
  select(-c(naics2, naics_match, lowcut, hicut))

saveRDS(pmtdb_match2, fs::path(PDINTERMEDIATE, "pmtdb_fpe.rds"))

```

```{r}
#| label: cleanup
rm(list = ls())
```

```{r stop_here, echo=FALSE}
knitr::knit_exit()
```

```{r}

# explore how much does fpe vary across counties within industry

look2 <- firms_per_estabs_match1 |> 
  mutate(esize_group = ntile(semp_estabs, 4), # divide the records into 4 establishment-size groups for each industry
         .by=c(snaics, stitle)) |> 
  select(area, snaics, stitle, esize_group, semp_estabs, sfirms_estabs) |> 
  summarise(
    semp_estabs_min = min(semp_estabs),
    semp_estabs_max = max(semp_estabs),
    sfirms_estabs_min = min(sfirms_estabs),
    sfirms_estabs_mdn = median(sfirms_estabs),
    sfirms_estabs_max = max(sfirms_estabs),
    .by=c(area, snaics, stitle, esize_group)) |> 
  arrange(area, snaics, esize_group) |> 
  # get industry-specific cutpoints for establishment-size groups (# of employees per establishment)
  mutate(lowcut = case_when(esize_group == 1 ~ 0,
                          esize_group > 1 ~ pmean(semp_estabs_min, lag(semp_estabs_max)),
                          .default = 0),
         hicut = case_when(esize_group == 4 ~ Inf,
                           esize_group < 4 ~ pmean(semp_estabs_max, lead(semp_estabs_min)),
                           .default = Inf)) |> 
  rename(naics_match = snaics) |> 
  arrange(naics_match, esize_group)

count(look2, esize_group)
count(look2, naics_match, stitle)

look2 |> 
  filter(naics_match == "51") |> 
  ggplot(aes(esize_group, sfirms_estabs_mdn, colour=area)) +
  geom_line()

```
