---
output: html_document
editor_options: 
  chunk_output_type: console
---

{{< include _setup.qmd >}}

# Get QCEW data

-   2023 is the latest full year as of 3/22/325
-   Get annual QCEW data for NY counties
-   Do minor cleaning
-   Put NYC and MTA markers on the file
-   Save as `qcewstco.rds` in the intermediate data folder
-   Save an MTA subset with just the latest year

## QCEW data from the New York State Department of Labor

-   [NYSDOL QCEW landing page](https://dol.ny.gov/quarterly-census-employment-and-wages)
-   [direct zip link](https://statistics.labor.ny.gov/qcew.zip)
-   [technical notes](https://dol.ny.gov/qcew-technical-notes)

The New York State Department of Labor (NYSDOL) provides a version of the BLS Quarterly Census of Employment and Wages (QCEW) that has quarterly county-level wage, establishment, and employment data in a format we need. The data are available from 2000 through a recent quarter. A new quarter and revisions to prior quarters generally are available about 6 months after end of a quarter.

```{r}
#| label: qcew-names
#| output: false

url <- "https://statistics.labor.ny.gov/qcew.zip"
fname <- fs::path_file(url)
zpath <- fs::path(PDRAW, fname)
zpath <- paste0(PDRAW, "/", fname)

```

## Download qcew data

Note: The latest qcew.zip file, which I downloaded once manually and once with the code below, appears to have been corrupted, at least as downloaded by my machine. I had to use an external utility (WinRar) to unzip it to a separate folder, then rezip it, after which the computer could read it.

Change eval: to true in chunk below to re-download data.

```{r}
#| label: download-qcew
#| eval: false

# Sys.setenv(R_DEFAULT_INTERNET_TIMEOUT = 400)  
options(timeout = max(400, getOption("timeout")))
download.file(url, zpath, mode="wb")

```

## Read, make minor adjustments, and save QCEW data

-   variable names --\> lower case
-   keep only state and county records - drop other kinds of areas and balance of state
-   add logical indicators for mta and nyc counties

```{r}
#| label: read-save-qcew
#| eval: false
#| output: false

# get names of the csv annual files and then read them

zfiles <- zip::zip_list(zpath)

csvfiles <- zfiles |> 
  filter(str_starts(filename, "qcew_annual_")) |> 
  pull(filename)

# read directly from the zip file
system.time(df <- vroom(purrr::map(csvfiles, \(x)  unz(zpath, x)))) # 2.8 secs
# obj_size(df)

df2 <- df |> 
  rename_with(str_to_lower)

# count(df2, areatype)
# count(df2, areatype, owner)

df3 <- df2 |> 
  filter(areatype %in% c("State", "County")) |> 
  mutate(areatype=str_to_lower(areatype)) |> 
  mutate(mta=area %in% constants$mtaregion$area,
         nyc=area %in% constants$nyc$area)

saveRDS(df3, fs::path(PDINTERMEDIATE, "qcewstco.rds"))

```

Save an MTA subset with just the latest year. Rename variables for consistency with rest of project.

```{r}
#| label: qcew-mta

qcew <- readRDS(fs::path(PDINTERMEDIATE, "qcewstco.rds"))
ns(qcew)
count(qcew, areatype)

qcewmta <- qcew |> 
  filter(year == max(year),
         areatype == "county",
         mta) |>
  mutate(naics_level=ifelse(naics=="00", 1, naics_level),
         area = str_remove(area, " County"),
         ownerf = factor(
           owner, 
           levels=c("Private", "Federal Government", "State Government", "Local Government", # details
                    "Total Private and Government", "Total Government"), # totals
           labels=1:6),
         ownerf = as.integer(as.character(ownerf))) |> 
    select(year, area, nyc, naics_level, naics, title, ownerf, owner, estabs=estab, emp=avgemp, payroll=totwage)
glimpse(qcewmta)
summary(qcewmta)
count(qcewmta, area)
count(qcewmta, naics_level, naics, title)
count(qcewmta, ownerf, owner)
qcewmta |> filter(payroll==max(payroll))

saveRDS(qcewmta, fs::path(PDINTERMEDIATE, "qcewmta.rds"))

```

```{r}
#| label: cleanup
rm(list = ls())
```

```{r stop_here, echo=FALSE}
knitr::knit_exit()
```
