# Actual and counterfactual -- State sales tax gain at 4% rate in entire state, MTA region, NYC and MTA Suburbs

**This analysis computes the STATE SALES TAX ONLY -- all calculations are done at the state 4% rate.**

It answers the question, How much did the state gain in different parts of the state?

There is some duplication between this and the page, *Actual and counterfactual -- State or NYC tax revenue gain* as I have not had time to streamline it yet.

Key dates:

-   January 15, 2019 State Tax Department notes informing vendors with no physical presence that they must register

-   June 1, 2019 Internet Marketplace Act applies to sales

-   March 15, 2020 (approximately) COVID-19 pandemic heats up, economy shuts down, store sales plummet (not shown in plots below), and online sales benefit

-   March 1, 2022 Sales tax year 2022-23 begins, with NAICS 4541 sales reclassified as NAICS 9999

```{r}
#| label: setup

library(tidyverse)
library(readxl)
library(vroom)
library(fs)
library(btools)
library(bggtools)
library(gt)
library(fable)
library(fabletools)
library(tsibble)
library(RcppRoll)

tprint <- 75  # default tibble print
options(tibble.print_max = tprint, tibble.print_min = tprint) # show up to tprint rows

ddir <- here::here("data")

fname <- "Taxable_Sales_And_Purchases_Quarterly_Data__Beginning_Sales_Tax_Year_2013-2014_dl20250128.csv"
fpath <- fs::path(ddir, fname)


mtasubcos <- c("Dutchess", "Nassau", "Orange", "Putnam", "Rockland", "Suffolk", "Westchester")
mtacos <- c("New York City", mtasubcos)

```

## Estimate actual values of internet taxable sales

-   Taxable sales and purchases for internet-related activity (NAICS 4541 in the 2017 NAICS) are not available after 2021-22 sales tax year
-   After that, the Tax Department put sales were put into NAICS 9999 unclassified. Eventually the Tax Department plans to put future such sales into their respective 2022 NAICS categories.
-   In the 2022-23+ period the former NAICS 4541 sales appear to constitute about 74% of NAICS 9999 sales, so we will use NAICS 9999 growth as a proxy for what NAICS 4541 would have grown.
-   Thus, estimate what actual NAICS 4541 sales would have been, for NYS and NYC, using two assumptions:
    1.  In the first year of missing data, 2022-23, assume internet taxable sales maintain their share of all industries, using the average share for the 2021-22 sales tax year (we do this because we don't yet have usable growth rates for industry 9999 because it ballooned in 2022-23 due to the reclassification)
    2.  After 2022-23 (when we now have a full year of the ballooned industry 9999 data), assume internet taxable sales grow at the same rate as taxable sales for industry 9999.

One improvement I could make is the transition from 2021-12 to 2022-03 assumes the average share of internet sales for the 2021-22 sales tax year. But there is some seasonality in the share. I could incorporate an estimate for that. I don't think it would have a material impact but would be a small nicety.

```{r}
#| label: estimate-actuals
#| output: false

df1 <- readRDS(path(ddir, "tsp_detail.rds"))
count(df1, mtacounty, uniname)

df2 <- df1 |> 
  filter(uniname %in% c("New York State", "MCTD", mtacos)) |> 
  mutate(groupname = case_when(
    uniname %in% mtasubcos ~ "MTA Suburban",
    uniname == "MCTD" ~ "MTA",
    .default = uniname))
glimpse(df2)
count(df2, mtacounty, uniname)
count(df2, groupname, uniname)

df3 <- df2 |> 
  summarise(txblsales = sum(txblsales, na.rm=TRUE),
            .by = c(stdate, naics, fnaics, groupname))
count(df3, groupname)

# get area sums of taxable sales over all industries
areasum <- df3 |> 
  summarise(n=n(),
            txblsales=sum(txblsales, na.rm = TRUE),
            .by=c(stdate, groupname)) |> 
  mutate(ind="all") |> 
  arrange(stdate)

# get area sums of taxable sales for industries of interest
inds <- df3 |> 
  filter(naics %in% c(4541, 9999)) |> 
  summarise(txblsales=sum(txblsales, na.rm = TRUE),
            .by=c(stdate, groupname, naics)) |> 
  mutate(ind=paste0("i", naics)) |> 
  select(-naics) |> 
  arrange(ind, stdate)

# combine all-industries and specific-industries data
# put in $ billions (divide by millions) and make wide
combo1 <- bind_rows(areasum, inds) |> 
  select(-n) |> 
  arrange(stdate) |> 
  mutate(txblsales = txblsales / 1e6) |> 
  pivot_wider(names_from=ind, values_from=txblsales)
combo1 |> filter(stdate >= "2021-03-01", stdate <= "2022-12-01") |> arrange(groupname, stdate)
# note that i4541 is missing beginning 2022-03-01 (start of 2022-23 sales tax year)
# and that i9999 leaps dramatically beginning 2022-03-01

# estimate i4541 using 2 assumptions:
#   1) in the first year of missing data, i4541 maintains its share of all industries, using the 
#      average share for the 2021-22 sales tax year -- so growth rates will be similar
#      note that this is a period for which IMA was in effect
#      also note that it is conservative - it assumes that i4541 grows at the same rate as all industries
#      we do this because we don't yet have meaningful i9999 growth values because of the data break
#   2) after that, it grows at the same rate as i9999, based on earlier analysis
combo2 <- combo1 |>
  arrange(groupname, stdate) |> 
  # get ma4 of the internet share of total, to use for 2022-23
  mutate(ishare = i4541 / all,
         isharema4 = RcppRoll::roll_mean(
           ishare, n=4,  align="right", fill=NA),
         .by=groupname) |> 
  # add i4541 estimated values for 2022-23 sales tax year assuming same growth as all industries (constant share)
  mutate(i4541est = i4541,
         i4541est = ifelse(stdate > "2021-12-01" & stdate <= "2022-12-01", 
                        all * isharema4[stdate == "2021-12-01"],
                        i4541),
         .by=groupname)  

fillmiss <- function(date, xmiss, xfull){
  # fill missing values in xmiss using growth rates of xfull
  # MUST BE SORTED BY DATE BEFORE CALLING
  xfullrate <- xfull / lag(xfull, 4)
  
  for(i in seq_along(date)){
    if(is.na(xmiss[i])){
      xmiss[i] <- xmiss[i-4] * xfullrate[i]
    }
  }
  xmiss
}

# add i4541 values beyond 2022-23 sales tax year assuming same growth as i9999
combo3 <- combo2 |> 
  arrange(stdate) |>  # djb 2025-03-13
  mutate(i4541est = fillmiss(stdate, i4541est, i9999),
         .by=groupname) |> 
  select(stdate, groupname, all, i4541, i9999, i4541est) |>
  pivot_longer(cols = -c(stdate, groupname),
               names_to = "indtype") |> 
  mutate(indf = factor(indtype,
                       levels=c("all", "i4541", "i9999", "i4541est"),
                       labels = c("all industries", "actual", "reclass", "estimated actual")))
count(combo3, indtype, indf)
glimpse(combo3)
combo3 |> filter(indtype=="i4541est") |> summarise(value=sum(value, na.rm = TRUE), .by=groupname)

saveRDS(combo3, fs::path(ddir, "internet_estimated_tsp_groups.rds"))

```

The results are shown below.

```{r}
#| label: show-actuals
#| output: true

df <- readRDS(fs::path(ddir, "internet_estimated_tsp_groups.rds"))

pdata <- df |> 
  filter(indtype %in% c("i4541", "i4541est")) |>
  filter(!(indtype=="i4541est" & stdate <= "2021-12-01")) |>
  filter(!is.na(value)) |>
  select(stdate, groupname, indtype, salestype=indf, value) |> 
  # mutate(uniname=factor(uniname, levels=c("New York State", "New York City"))) |> # for facet order
  arrange(groupname, desc(salestype), stdate)

pdata |>
  ggplot(aes(x=stdate, y=value, colour=salestype)) +
  geom_line(linewidth=1) +
  geom_point(size=1.25) +
  scale_colour_manual(values=c("actual"="darkgreen", "estimated actual"="blue")) +
  geom_vline(xintercept = as.Date("2019-06-01"), colour="red") +
  geom_vline(xintercept = as.Date("2020-03-15"), colour="black", linetype="dashed") +
  geom_vline(xintercept = as.Date("2022-03-01"), colour="blue") +  
  facet_wrap(~groupname, scales="free_y", ncol=1) +
  ggtitle("Actual and estimated actual quarterly internet taxable sales (NAICS 4541, 2017 basis)",
          subtitle="Internet Marketplace Act effective 2019-06-01; COVID-19 start ~2020-03-15; Sales reclassification 2022-03-01") +
  scale_x_date(name=NULL, breaks = scales::date_breaks("1 year"), date_labels = "%Y") +
  scale_y_continuous(name="Taxable sales ($ billions)", labels=scales::dollar) +
  theme_bw() +
  legend_notitle

```

## Estimate counterfactual values of internet taxable sales

Extrapolate internet sales for 2 years from 2019-06-01 to 2021-06-01 based on prior 2 year growth rate then constant share of est actual from there forward

```{r}
#| label: estimate-counterfactual-internet-sales_extrap
#| output: false

tsales1 <- readRDS(fs::path(ddir, "internet_estimated_tsp_groups.rds"))
head(tsales1)
summary(tsales1)
count(tsales1, indtype, indf)

tsales2 <- tsales1 |> 
  filter(indtype %in% c("all", "i4541", "i4541est")) |>
  select(stdate, groupname, indtype, value) |> 
  pivot_wider(names_from=indtype, values_from=value) |>
  arrange(groupname, stdate) |> # djb 2025-03-11
  mutate(i4541counter=ifelse(stdate < as.Date("2019-06-01"), i4541est, NA_real_)) |> 
  mutate(ratio2ya = i4541est / lag(i4541est, 8), .by=groupname) |> 
  mutate(i4541counter = case_when(stdate >= as.Date("2019-06-01") & 
                                    stdate <= as.Date("2021-06-01") ~ lag(i4541est, 8) * ratio2ya[stdate=="2019-03-01"],
                                  stdate > as.Date("2021-06-01") ~ NA_real_,
                                  .default = i4541counter),
         .by=groupname) |> 
  mutate(share=i4541counter / i4541est,
         i4541counter=ifelse(stdate >= "2021-09-01",
                             i4541est * share[stdate=="2021-06-01"],
                             i4541counter),
         .by=groupname) # djb 3/14/2025

# tsales2 |> 
#   ggplot(aes(x=stdate, y=i4541counter, color=uniname)) +
#   geom_line() +
#   geom_point()

tsales3 <- tsales2 |> 
  select(stdate, groupname, all, starts_with("i4541")) |> 
  pivot_longer(cols=-c(stdate, groupname), names_to="indtype", values_to="value") |> 
  mutate(indf = factor(indtype,
                       levels=c("all", "i4541", "i4541est", "i4541counter"),
                       labels = c("all industries", "actual", "estimated actual", "counterfactual")))

```

```{r}
#| label: show-counter
#| output: true

pdata <- tsales3 |> 
  filter(str_detect(indtype, "i4541")) |> 
  rename(salestype=indf)
# count(pdata, salestype)

pdata |>
  # filter(!(indtype!="i4541" & stdate < as.Date("2019-06-01"))) |> 
  filter((indtype == "i4541" & stdate < as.Date("2022-03-01")) | # actual
           (indtype == "i4541est" & stdate >= as.Date("2022-03-01")) |  # estimated
           (indtype == "i4541counter" & stdate >= as.Date("2019-06-01"))) |>
  ggplot(aes(x=stdate, y=value, colour=salestype)) +
  geom_line(linewidth=1) +
  geom_point(size=1.25) +
  scale_colour_manual(values=c("actual"="darkgreen", "estimated actual"="blue", "counterfactual"= "grey")) +
  geom_vline(xintercept = as.Date("2019-06-01"), colour="red") +
  geom_vline(xintercept = as.Date("2020-03-15"), colour="black", linetype="dashed") +
  geom_vline(xintercept = as.Date("2022-03-01"), colour="blue") +  
  facet_wrap(~groupname, scales="free_y", ncol=1) +
  ggtitle("Actual and estimated actual quarterly internet taxable sales (NAICS 4541, 2017 basis)",
          subtitle="Internet Marketplace Act effective 2019-06-01; COVID-19 start ~2020-03-15; Sales reclassification 2022-03-01") +
  scale_x_date(name=NULL, breaks = scales::date_breaks("1 year"), date_labels = "%Y") +
  scale_y_continuous(name="Taxable sales ($ millions)", labels=scales::dollar) +
  theme_bw() +
  legend_notitle

```

Here are rough estimates of tax revenue gains by area

```{r}
#| label: show-taxes
#| output: true

# count(tsales3, groupname)

ima <- tsales3 |> 
  filter(indtype %in% c("i4541est", "i4541counter")) |>
  select(-indf) |> 
  pivot_wider(names_from=indtype, values_from=value) |> 
  mutate(gain=i4541est - i4541counter,
         taxgain=.04 * gain) |> 
  arrange(groupname, stdate) |> # djb 2025-03-11
  mutate(taxgain4=RcppRoll::roll_sumr(taxgain, n=4), .by=groupname)

ima |> 
  filter(year(stdate) >= 2019) |> 
  gt() |> 
  tab_header(title="Estimated NY STATE tax revenue gains from Internet Marketplace Act, $ millions by region",
             subtitle="4% state tax rate used for all calculations") |> 
  cols_label(stdate="Sales tax quarter beginning:",
             groupname="Region",
             i4541est="Estimated actual taxable sales",
             i4541counter="Counterfactual taxable sales",
             gain="Estimated taxable sales gain",
             taxgain="Estimated tax revenue gain",
             taxgain4="4-quarter sum of tax revenue gain") |> 
  fmt_number(columns=c(i4541est, i4541counter, gain), decimals=0) |>
  fmt_number(columns=c(taxgain, taxgain4), decimals=0)

```

And here are just the gains, all regions on one line

```{r}
#| label: show-taxgains
#| output: true

ima |> 
  filter(year(stdate) >= 2019) |> 
  select(stdate, groupname, taxgain4) |> 
  pivot_wider(names_from = groupname, values_from = taxgain4) |> 
  relocate(`New York State`, .before = MTA) |> 
  gt() |> 
  tab_header(title="Estimated 4-quarter sums of NY STATE tax revenue gains from Internet Marketplace Act, $ millions by region",
             subtitle="4% state tax rate used for all calculations") |> 
  cols_label(stdate="Sales tax quarter beginning:") |> 
  fmt_number(columns=-stdate, decimals = 1)


```

```{r}
#| label: stoprun
#| echo: false
knitr::knit_exit()  # Stops rendering here
```
