# Actual and counterfactual

Key dates:

-   January 15, 2019 State Tax Department notes informing vendors with no physical presence that they must register

-   June 1, 2019 Internet Marketplace Act applies to sales

-   March 15, 2020 (approximately) COVID-19 pandemic heats up, economy shuts down, store sales plummet (not shown in plots below), and online sales benefit

-   March 1, 2022 Sales tax year 2022-23 begins, with NAICS 4541 sales reclassified as NAICS 9999

```{r}
#| label: setup

# source(here::here("r", "libraries.r"))
# source(here::here("r", "libraries_ts.r"))
# source(here::here("r", "constants.r"))
# source(here::here("r", "functions.r"))

library(tidyverse)
library(readxl)
library(vroom)
library(fs)
library(btools)
library(bggtools)
library(gt)
library(fable)
library(fabletools)
library(tsibble)
library(RcppRoll)

tprint <- 75  # default tibble print
options(tibble.print_max = tprint, tibble.print_min = tprint) # show up to tprint rows

ddir <- here::here("data")

fname <- "Taxable_Sales_And_Purchases_Quarterly_Data__Beginning_Sales_Tax_Year_2013-2014_dl20250128.csv"
fpath <- fs::path(ddir, fname)

```

## Estimate actual values of internet taxable sales

-   Taxable sales and purchases for internet-related activity (NAICS 4541 in the 2017 NAICS) are not available after 2021-22 sales tax year
-   After that, the Tax Department put sales were put into NAICS 9999 unclassified. Eventually the Tax Department plans to put future such sales into their respective 2022 NAICS categories.
-   In the 2022-23+ period the former NAICS 4541 sales appear to constitute about 74% of NAICS 9999 sales, so we will use NAICS 9999 growth as a proxy for what NAICS 4541 would have grown.
-   Thus, estimate what actual NAICS 4541 sales would have been, for NYS and NYC, using two assumptions:
    1.  In the first year of missing data, 2022-23, assume internet taxable sales maintain their share of all industries, using the average share for the 2021-22 sales tax year (we do this because we don't yet have usable growth rates for industry 9999 because it ballooned in 2022-23 due to the reclassification)
    2.  After 2022-23 (when we now have a full year of the ballooned industry 9999 data), assume internet taxable sales grow at the same rate as taxable sales for industry 9999.

One improvement I could make is the transition from 2021-12 to 2022-03 assumes the average share of internet sales for the 2021-22 sales tax year. But there is some seasonality in the share. I could incorporate an estimate for that. I don't think it would have a material impact but would be a small nicety.

```{r}
#| label: estimate-actuals
#| output: false

df1 <- readRDS(path(ddir, "tsp_detail.rds"))

# uname <- "New York State"
# 
# uname <- "New York City"

# get state and city sums of taxable sales over all industries
stsum <- df1 |> 
  filter(uniname %in% c("New York State", "New York City")) |>
  summarise(n=n(),
            txblsales=sum(txblsales, na.rm = TRUE),
            .by=c(stdate, uniname)) |> 
  mutate(ind="all") |> 
  arrange(stdate)

# get state and city sums of taxable sales for industries of interest
inds <- df1 |> 
  filter(uniname %in% c("New York State", "New York City"),
         naics %in% c(4541, 9999)) |> 
  summarise(txblsales=sum(txblsales, na.rm = TRUE),
            .by=c(stdate, uniname, naics)) |> 
  mutate(ind=paste0("i", naics)) |> 
  select(-naics) |> 
  arrange(ind, stdate)

# combine all-industries and specific-industries data
# put in $ billions (divide by millions) and make wide
combo1 <- bind_rows(stsum, inds) |> 
  select(-n) |> 
  arrange(stdate) |> 
  mutate(txblsales = txblsales / 1e6) |> 
  pivot_wider(names_from=ind, values_from=txblsales)
combo1 |> tail(24)
# note that i4541 is missing beginning 2022-03-01 (start of 2022-23 sales tax year)
# and that i9999 leaps dramatically beginning 2022-03-01

# estimate i4541 using 2 assumptions:
#   1) in the first year of missing data, i4541 maintains its share of all industries, using the 
#      average share for the 2021-22 sales tax year -- so growth rates will be similar
#      note that this is a period for which IMA was in effect
#      also note that it is conservative - it assumes that i4541 grows at the same rate as all industries
#      we do this because we don't yet have meaningful i9999 growth values because of the data break
#   2) after that, it grows at the same rate as i9999, based on earlier analysis
combo2 <- combo1 |>
  arrange(uniname, stdate) |> 
  # get ma4 of the internet share of total, to use for 2022-23
  mutate(ishare = i4541 / all,
         isharema4 = RcppRoll::roll_mean(
           ishare, n=4,  align="right", fill=NA),
         .by=uniname) |> 
  # add i4541 estimated values for 2022-23 sales tax year assuming same growth as all industries (constant share)
  mutate(i4541est = i4541,
         i4541est = ifelse(stdate > "2021-12-01" & stdate <= "2022-12-01", 
                        all * isharema4[stdate == "2021-12-01"],
                        i4541),
         .by=uniname)  

fillmiss <- function(date, xmiss, xfull){
  # fill missing values in xmiss using growth rates of xfull
  xfullrate <- xfull / lag(xfull, 4)
  
  for(i in seq_along(date)){
    if(is.na(xmiss[i])){
      xmiss[i] <- xmiss[i-4] * xfullrate[i]
    }
  }
  xmiss
}

# add i4541 values beyond 2022-23 sales tax year assuming same growth as i9999
combo3 <- combo2 |> 
  mutate(i4541est = fillmiss(stdate, i4541est, i9999),
         .by=uniname) |> 
  select(stdate, uniname, all, i4541, i9999, i4541est) |>
  pivot_longer(cols = -c(stdate, uniname),
               names_to = "indtype") |> 
  mutate(indf = factor(indtype,
                       levels=c("all", "i4541", "i9999", "i4541est"),
                       labels = c("all industries", "actual", "reclass", "estimated actual")))
count(combo3, indtype, indf)
glimpse(combo3)

saveRDS(combo3, fs::path(ddir, "internet_estimated_tsp.rds"))

# |> 
#   mutate(across(c(all, i4541, i9999, i4541est), list(pch = \(x) x / lag(x, 4) - 1)),
#          .by=uniname) |> 
#   mutate(ishareest = i4541 / all,
#          isharema4est = RcppRoll::roll_mean(
#            ishareest, n=4,  align="right", fill=NA),
#          salestype=ifelse(stdate > "2021-12-01", "estimated actual", "actual"),
#          .by=uniname)

# combo3 |> 
#   filter(!is.na(isharema4)) |>
#   select(stdate, uniname, ishare, isharema4) |> 
#   pivot_longer(cols=c(ishare, isharema4), names_to="isharetype", values_to="ishare") |>
#   ggplot(aes(x=stdate, y=ishare, color=isharetype)) +
#   geom_line() +
#   geom_point() +
#   geom_vline(xintercept = as.Date("2019-06-01")) +
#   geom_vline(xintercept = as.Date("2022-03-01"), colour="blue") +
#   facet_wrap(~uniname, scales="free_y")

# combo3 |> 
#   select(stdate, uniname, i4541) |> 
#   ggplot(aes(x=stdate, y=i4541)) +
#   geom_line() +
#   geom_point() +
#   geom_vline(xintercept = as.Date("2019-06-01")) +
#   geom_vline(xintercept = as.Date("2022-03-01"), colour="blue") +  
#   facet_wrap(~uniname, scales="free_y")

```

The results are shown below.

```{r}
#| label: show-actuals
#| output: true

df <- readRDS(fs::path(ddir, "internet_estimated_tsp.rds"))

pdata <- df |> 
  filter(indtype %in% c("i4541", "i4541est")) |>
  filter(!(indtype=="i4541est" & stdate <= "2021-12-01")) |>
  filter(!is.na(value)) |>
  select(stdate, uniname, indtype, salestype=indf, value) |> 
  mutate(uniname=factor(uniname, levels=c("New York State", "New York City"))) |> # for facet order
  arrange(uniname, desc(salestype), stdate)

pdata |>
  ggplot(aes(x=stdate, y=value, colour=salestype)) +
  geom_line(linewidth=1) +
  geom_point(size=1.25) +
  scale_colour_manual(values=c("actual"="darkgreen", "estimated actual"="blue")) +
  geom_vline(xintercept = as.Date("2019-06-01"), colour="red") +
  geom_vline(xintercept = as.Date("2020-03-15"), colour="black", linetype="dashed") +
  geom_vline(xintercept = as.Date("2022-03-01"), colour="blue") +  
  facet_wrap(~uniname, scales="free_y", ncol=1) +
  ggtitle("Actual and estimated actual quarterly internet taxable sales (NAICS 4541, 2017 basis)",
          subtitle="Internet Marketplace Act effective 2019-06-01; COVID-19 start ~2020-03-15; Sales reclassification 2022-03-01") +
  scale_x_date(name=NULL, breaks = scales::date_breaks("1 year"), date_labels = "%Y") +
  scale_y_continuous(name="Taxable sales ($ billions)", labels=scales::dollar) +
  theme_bw() +
  legend_notitle

```

## Estimate counterfactual values of internet taxable sales

Use a simple seasonal ARIMA model to forecast counterfactual industry 4541 taxable sales, based on pre-IMA data. Estimate it in log form. (One model for NYS and one for NYC.) Use the model to forecast counterfactual sales for the post-IMA period.

This could be improved by using a regression with ARIMA errors model for wages.

```{r}
#| label: estimate-counterfactual-internet-sales
#| output: false

tsales1 <- readRDS(fs::path(ddir, "internet_estimated_tsp.rds"))
head(tsales1)
summary(tsales1)

datesdf <- tibble(stdate=seq.Date(from=as.Date("2013-03-01"), to=as.Date("2030-12-01"), by="quarter")) |> 
  mutate(yqdate=yearquarter(stdate))

tsales2 <- tsales1 |> 
  filter(indtype=="i4541est", stdate < as.Date("2019-06-01")) |>
  mutate(yqdate=yearquarter(stdate)) |> 
  arrange(uniname, yqdate) |> 
  as_tsibble(key=uniname, index = yqdate)

tsales_fit <- tsales2 |>
  model(arima = ARIMA(log(value) ~ pdq(d=1) + PDQ(D=1, period=4)))
  # model(arima = ARIMA(log(value) ~ trend() + pdq(d=1) + PDQ(D=1, period=4)))
  # model(arima = ARIMA(log(value) ~ 1 + pdq(d=1) + PDQ(D=1, period=4)))

tsales_fit
glance(tsales_fit)
tidy(tsales_fit)
report(tsales_fit |> select(arima) |> filter(uniname == "New York State"))
report(tsales_fit |> select(arima) |> filter(uniname == "New York City"))

tsales_fc <- tsales_fit |> 
  forecast(h = 22)
tsales_fc

tsales_fc2 <- tsales_fc |> 
  left_join(datesdf)

# put the data together

result <- tsales2 |> 
  as_tibble() |> 
  bind_rows(tsales_fc2 |> 
              as_tibble() |> 
              select(stdate, uniname, value=.mean)) |>
  mutate(indtype="i4541counter",
         indf="counterfactual") 

combo <- bind_rows(
  tsales1,
  result)
  

# Generate forecasts
# wages_forecast <- wages_fit %>%
#   forecast(h = "4 years") 
# 
# addon <- wages_forecast |> 
#   as_tibble() |> 
#   select(date, wages=.mean) |> 
#   mutate(type="forecast")
# 
# result <- bind_rows(
#   wages_data |> mutate(type="actual"),
#   addon)
# 
# result |> 
#   ggplot(aes(date, wages, color=type)) +
#   geom_line() +
#   geom_point()
# 
# wages_forecast |> 
#   mutate(intervals = hilo(wages, 95)) |> 
#   unpack_hilo(intervals)

```

Variant

extrapolate int sales for 2 years from 2019-06-01 to 2021-06-01 based on prior 2 year growth rate then constant share of est actual from there forward

```{r}
#| label: estimate-counterfactual-internet-sales_extrap
#| output: false

tsales1 <- readRDS(fs::path(ddir, "internet_estimated_tsp.rds"))
head(tsales1)
summary(tsales1)
count(tsales1, indtype, indf)

tsales2 <- tsales1 |> 
  filter(indtype %in% c("all", "i4541", "i4541est")) |>
  select(stdate, uniname, indtype, value) |> 
  mutate(uniname=factor(uniname, levels=c("New York State", "New York City"))) |> 
  pivot_wider(names_from=indtype, values_from=value) |>
  mutate(i4541counter=ifelse(stdate < as.Date("2019-06-01"), i4541est, NA_real_),) |> 
  mutate(ratio2ya = i4541est / lag(i4541est, 8), .by=uniname) |> 
  mutate(i4541counter = case_when(stdate >= as.Date("2019-06-01") & 
                                    stdate <= as.Date("2021-06-01") ~ lag(i4541est, 8) * ratio2ya[stdate=="2019-03-01"],
                                  stdate > as.Date("2021-06-01") ~ NA_real_,
                                  .default = i4541counter),
         .by=uniname) |> 
  mutate(share=i4541counter / i4541est,
         i4541counter=ifelse(stdate >= "2021-09-01",
                             i4541est * share[stdate=="2021-06-01"],
                             i4541counter)) 

# tsales2 |> 
#   ggplot(aes(x=stdate, y=i4541counter, color=uniname)) +
#   geom_line() +
#   geom_point()

tsales3 <- tsales2 |> 
  select(stdate, uniname, all, starts_with("i4541")) |> 
  pivot_longer(cols=-c(stdate, uniname), names_to="indtype", values_to="value") |> 
  mutate(indf = factor(indtype,
                       levels=c("all", "i4541", "i4541est", "i4541counter"),
                       labels = c("all industries", "actual", "estimated actual", "counterfactual")))

```

```{r}
#| label: show-counter
#| output: true

pdata <- tsales3 |> 
  filter(str_detect(indtype, "i4541")) |> 
  rename(salestype=indf)
# count(pdata, salestype)

pdata |>
  # filter(!(indtype!="i4541" & stdate < as.Date("2019-06-01"))) |> 
  filter((indtype == "i4541" & stdate < as.Date("2022-03-01")) | # actual
           (indtype == "i4541est" & stdate >= as.Date("2022-03-01")) |  # estimated
           (indtype == "i4541counter" & stdate >= as.Date("2019-06-01"))) |>
  ggplot(aes(x=stdate, y=value, colour=salestype)) +
  geom_line(linewidth=1) +
  geom_point(size=1.25) +
  scale_colour_manual(values=c("actual"="darkgreen", "estimated actual"="blue", "counterfactual"= "grey")) +
  geom_vline(xintercept = as.Date("2019-06-01"), colour="red") +
  geom_vline(xintercept = as.Date("2020-03-15"), colour="black", linetype="dashed") +
  geom_vline(xintercept = as.Date("2022-03-01"), colour="blue") +  
  facet_wrap(~uniname, scales="free_y", ncol=1) +
  ggtitle("Actual and estimated actual quarterly internet taxable sales (NAICS 4541, 2017 basis)",
          subtitle="Internet Marketplace Act effective 2019-06-01; COVID-19 start ~2020-03-15; Sales reclassification 2022-03-01") +
  scale_x_date(name=NULL, breaks = scales::date_breaks("1 year"), date_labels = "%Y") +
  scale_y_continuous(name="Taxable sales ($ millions)", labels=scales::dollar) +
  theme_bw() +
  legend_notitle

```

Here are rough estimates of tax revenue gains by NYS and NYC

```{r}
#| label: show-taxes
#| output: true

ima <- tsales3 |> 
  filter(indtype %in% c("i4541est", "i4541counter")) |>
  select(-indf) |> 
  pivot_wider(names_from=indtype, values_from=value) |> 
  mutate(gain=i4541est - i4541counter,
         taxgain=ifelse(uniname=="New York State",
                    .04 * gain,
                    .045 * gain)) |> 
  mutate(taxgain4=RcppRoll::roll_sumr(taxgain, n=4), .by=uniname)

ima |> 
  filter(year(stdate) >= 2019) |> 
  gt() |> 
  tab_header(title="Estimated tax revenue gains from Internet Marketplace Act, $ millions",
             subtitle="Assuming 4% state and 4.5% city tax rates") |> 
  cols_label(stdate="Sales tax quarter beginning:",
             uniname="Jurisdiction",
             i4541est="Estimated actual taxable sales",
             i4541counter="Counterfactual taxable sales",
             gain="Estimated taxable sales gain",
             taxgain="Estimated tax revenue gain",
             taxgain4="4-quarter sum of tax revenue gain") |> 
  fmt_number(columns=c(i4541est, i4541counter, gain), decimals=0) |>
  fmt_number(columns=c(taxgain, taxgain4), decimals=0)




```

```{r}
#| echo: false
knitr::knit_exit()  # Stops rendering here
```

```{r}
#| label: show-counter2
#| output: false
#| eval: false

pdata <- combo |> 
  filter(str_detect(indtype, "i4541")) |> 
  rename(salestype=indf)

pdata |>
  ggplot(aes(x=stdate, y=value, colour=salestype)) +
  geom_line(linewidth=1) +
  geom_point(size=1.25) +
  scale_colour_manual(values=c("actual"="darkgreen", "estimated actual"="blue", "counterfactual"= "grey")) +
  geom_vline(xintercept = as.Date("2019-06-01"), colour="red") +
  geom_vline(xintercept = as.Date("2020-03-15"), colour="black", linetype="dashed") +
  geom_vline(xintercept = as.Date("2022-03-01"), colour="blue") +  
  facet_wrap(~uniname, scales="free_y", ncol=1) +
  ggtitle("Actual and estimated actual quarterly internet taxable sales (NAICS 4541, 2017 basis)",
          subtitle="Internet Marketplace Act effective 2019-06-01; COVID-19 start ~2020-03-15; Sales reclassification 2022-03-01") +
  scale_x_date(name=NULL, breaks = scales::date_breaks("1 year"), date_labels = "%Y") +
  scale_y_continuous(name="Taxable sales ($ billions)", labels=scales::dollar) +
  theme_bw() +
  legend_notitle

```

```{r}
#| output: false
#| eval: false


tsales1 <- readRDS(fs::path(ddir, "internet_estimated_tsp.rds"))
head(tsales1)
summary(tsales1)
count(tsales1, indtype, indf)

tsales2 <- tsales1 |> 
  filter(indtype %in% c("all", "i4541", "i4541est")) |>
  select(stdate, uniname, indtype, value) |> 
  mutate(uniname=factor(uniname, levels=c("New York State", "New York City"))) |> 
  pivot_wider(names_from=indtype, values_from=value) |>
  mutate(other=all - i4541est,
         share_other = i4541est / other)

tsales2 |> 
  ggplot(aes(x=stdate, y=share_other, color=uniname)) +
  geom_line() +
  geom_point()


tsales3 <- tsales2 |> 
  mutate(i4541counter=ifelse(stdate < as.Date("2019-06-01"), i4541est, NA_real_)) |> 
  mutate(dshare=share_other - lag(share_other, 4), .by=uniname,
         qtr=quarter(stdate),
         year=year(stdate)) |> 
  mutate(fcshare=share_other[stdate==as.Date("2019-06-01")],
         .by=uniname) |> 
  mutate(i4541counter=ifelse(is.na(i4541counter), 
                             other * fcshare,
                             i4541counter))

tsales3
tsales4 <- tsales3 |> 
  select(stdate, uniname, all, starts_with("i4541")) |> 
  pivot_longer(cols=-c(stdate, uniname), names_to="indtype", values_to="value") |> 
  mutate(indf = factor(indtype,
                       levels=c("all", "i4541", "i4541est", "i4541counter"),
                       labels = c("all industries", "actual", "estimated actual", "counterfactual")))

```
